---
title: "NSW Road Crash Severity Prediction"
subtitle: "STAT5003 — Group 07, Workshop 12"
author: "Sanskar | Robert | Haipeng | Chen | Yihan"
date: "May 2026"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.align = "center", out.width = "90%")

library(ggplot2)
library(dplyr)
library(readr)
library(forcats)
library(scales)
library(knitr)
library(kableExtra)
library(tidyr)
library(flextable)
```

```{r load-data, include=FALSE}
df_raw <- read_csv("nsw_road_crash_merged_2020_2024.csv", show_col_types = FALSE)

df_raw <- df_raw %>%
  rename(severity = `Degree of crash`) %>%
  mutate(severity = factor(severity,
    levels = c("Fatal", "Injury", "Non-casualty (towaway)"),
    labels = c("Fatal", "Injury", "Non-casualty")))
```

## Research Question & Motivation

**Research Question:**

> Given environmental conditions (weather, lighting, road surface, speed limit) 
> and traffic unit characteristics (TU type, manoeuvre, number of units), 
> can we predict NSW road crash severity **(Fatal / Injury / Non-casualty)**?
> — A three-class classification problem.

**Motivation:**

- NSW recorded **340 road fatalities in 2024**, over two-thirds in regional areas *(TfNSW, 2025)*
- Fatal crashes represent only **~1.4%** of all crashes → severe class imbalance
- A predictive model has direct value for:

| Stakeholder | Value |
|-------------|-------|
| **Traffic Authorities** | Targeted speed and road interventions |
| **Emergency Responders** | Faster, evidence-based resource allocation |
| **Policymakers & Insurers** | Precise policy and pricing decisions |

## Dataset Overview

```{r dataset-overview-robert, echo=FALSE, message=FALSE, warning=FALSE}
total_records <- nrow(df_raw)
total_vars    <- ncol(df_raw)
class_counts  <- table(df_raw$severity)
class_props   <- round(prop.table(class_counts) * 100, 1)

overview <- data.frame(
  Item = c(
    "Time Period",
    "Data Sources",
    "Merge Key",
    "Total Records",
    "Total Variables",
    "Target Variable",
    "Fatal",
    "Injury",
    "Non-casualty"
  ),
  Detail = c(
    "2020 – 2024",
    "CRASH.xlsx + TRAFFIC UNIT.xlsx (TfNSW via data.gov.au)",
    "Crash ID (left join)",
    format(total_records, big.mark = ","),
    as.character(total_vars),
    "Degree of Crash (3-class classification)",
    paste0(format(class_counts["Fatal"], big.mark=","), 
           " (", class_props["Fatal"], "%)"),
    paste0(format(class_counts["Injury"], big.mark=","), 
           " (", class_props["Injury"], "%)"),
    paste0(format(class_counts["Non-casualty"], big.mark=","), 
           " (", class_props["Non-casualty"], "%)")
  )
)

flextable(overview) |>
  set_header_labels(Item = "Item", Detail = "Detail") |>
  theme_vanilla() |>
  bold(j = 1) |>
  bold(part = "header") |>
  bg(i = c(2, 4, 6), bg = "#f5f5f5") |>
  color(i = 7, j = 2, color = "#d62728") |>
  color(i = 8, j = 2, color = "#ff7f0e") |>
  color(i = 9, j = 2, color = "#2ca02c") |>
  fontsize(size = 16, part = "all") |>    
  fontsize(size = 18, part = "header") |> 
  padding(padding = 8, part = "all") |> 
  set_table_properties(width = 1) |>   
  autofit()
```

## Class Imbalance

```{r class-imbalance-robert, echo=FALSE, message=FALSE, warning=FALSE, fig.width=7, fig.height=4}

class_dist <- df_raw %>%
  count(severity) %>%
  mutate(
    proportion = round(n / sum(n) * 100, 1),
    label = paste0(proportion, "%\n(n=", format(n, big.mark=","), ")")
  )

ggplot(class_dist,
       aes(x = severity, y = proportion, fill = severity)) +
  geom_bar(stat = "identity", width = 0.6) +
  geom_text(
    aes(label = label),
    vjust = -0.3,
    size  = 4,
    fontface = "bold"
  ) +
  scale_fill_manual(values = c(
    "Fatal"        = "#d62728",
    "Injury"       = "#ff7f0e",
    "Non-casualty" = "#2ca02c"
  )) +
  scale_y_continuous(limits = c(0, 80)) +
  labs(
    title    = "Class Distribution of Crash Severity (2020–2024)",
    subtitle = "Fatal crashes are severely underrepresented",
    x = "Severity Class",
    y = "Proportion (%)"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

> ⚠️ **Fatal crashes (~1.4%) are severely underrepresented**  
> → SMOTE applied **inside training fold only**  
> → Primary metrics: **Macro-F1** and **Fatal Recall (≥ 0.50)**
## EDA & Data Understanding: Dataset Overview

<small style="color:#2E74B5">**Rubric criterion:** EDA & Data Understanding  \|  2 pts</small>

```{r slide-dataset}
# TODO: [member name] — dataset summary table (source, records, variables, target, date range)
```

## EDA & Data Understanding: Class Imbalance

<small style="color:#2E74B5">**Rubric criterion:** EDA & Data Understanding  \|  2 pts</small>

```{r slide-imbalance, fig.height=3.5}
# TODO: [member name] — bar chart of Fatal / Injury / Non-casualty proportions
```

## EDA & Data Understanding: Speed vs Severity

<small style="color:#2E74B5">**Rubric criterion:** EDA & Data Understanding  \|  2 pts</small>

```{r slide-speed, fig.height=3.5}
# TODO: [member name] — stacked bar chart of severity proportion by speed limit band
```

## EDA & Data Understanding: High-Risk Scenarios

<small style="color:#2E74B5">**Rubric criterion:** EDA & Data Understanding  \|  2 pts</small>

```{r slide-highrisk, fig.height=3.5}
# TODO: [member name] — fatal rate by speed band and lighting condition
```

## EDA & Data Understanding: Temporal Patterns

<small style="color:#2E74B5">**Rubric criterion:** EDA & Data Understanding  \|  2 pts</small>

```{r slide-temporal, fig.height=3.5}
# TODO: [member name] — crash frequency heatmap by day of week and time interval
```

## Data Preprocessing: Pipeline

<small style="color:#2E74B5">**Rubric criterion:** Data Preprocessing  \|  2 pts</small>

**Steps 1–5: Cleaning & Feature Engineering**

1.  **TU Integration** — aggregated traffic-unit flags per crash (`has_motorcycle`, `has_heavy_vehicle`, `has_pedestrian`, `has_cyclist`) using `group_by + summarise + left_join`
2.  **Feature removal** — dropped leakage columns (post-crash injury counts), identifiers, and variables with \>75% missing values; dropped `Degree of crash - detailed` (redundant with target)
3.  **Imputation** — removed rows \>50% missing; categorical → `"Unknown"`; numeric → median
4.  **Feature engineering** — `is_weekend`, `is_rush_hour`, `is_night`, `is_wet`, speed bands — all motivated by EDA findings
5.  **Outlier handling** — NSW geographic bounds filter; winsorise `Distance` at 99th percentile

**Steps 6–9: Modelling Setup**

6.  **Encoding** — trees use native R `factor`; KNN/LR use one-hot; rare LGA levels → `"Other"`
7.  **Scaling** — Z-score standardisation within fold for KNN; trees are scale-invariant
8.  **Class imbalance** — inverse-frequency class weights (all models); SMOTE applied **inside training fold only**
9.  **Split** — 2024 data sealed as temporal holdout; 80/20 stratified split on 2020–2023; 5-fold CV

## Data Preprocessing: Feature Engineering & Class Balancing

<small style="color:#2E74B5">**Rubric criterion:** Data Preprocessing  \|  2 pts</small>

**Engineered features — all motivated by EDA**

| Feature | Derivation | EDA motivation |
|------------------------|------------------------|------------------------|
| `speed_band` | Cut `Speed limit` into ≤50 / 60–80 / 90–100 / 110 | Fatal rate rises sharply above 100 km/h |
| `is_night` | `Natural lighting` ∈ {Darkness, Dawn, Dusk} | Dark conditions double fatal risk |
| `is_rush_hour` | `Two-hour intervals` ∈ {7–9, 16–18} | Commute peaks dominate crash volume |
| `is_weekend` | `Day of week of crash` ∈ Saturday, Sunday | Weekend crashes differ in severity profile |
| `is_wet` | `Surface condition` contains "Wet" | Wet roads linked to skidding fatalities |
| `has_motorcycle` | TU-level aggregated flag | Motorcyclists over-represented in fatals |
| `has_pedestrian` | TU-level aggregated flag | Pedestrian crashes nearly always severe |

**Handling class imbalance**

-   Raw Fatal share \~1.4% → accuracy-maximising models ignore Fatal entirely
-   Strategy: **inverse-frequency class weights** (all models) + **SMOTE** inside training fold (LR, KNN)
-   SMOTE raises synthetic Fatal share to \~10% in training only — test set remains unmodified

## Model Fitting: Approach & Setup

<small style="color:#2E74B5">**Rubric criterion:** Model Fitting  \|  2 pts</small>

```{r model-overview}
model_tbl <- tibble::tibble(
  Model              = c("Multinomial Logistic Regression", "Decision Tree",
                         "Random Forest", "XGBoost",
                         "K-Nearest Neighbours", "Naive Bayes"),
  `R package`        = c("nnet", "rpart", "randomForest", "xgboost", "kknn", "e1071"),
  `Why chosen`       = c(
    "Interpretable baseline; log-odds show predictor direction",
    "Fully interpretable decision rules; native categorical handling",
    "Ensemble; robust to noise; provides variable importance",
    "Gradient boosting; typically strongest on tabular data",
    "Non-parametric; captures local structure without assumptions",
    "Probabilistic baseline; reveals feature independence violations"
  ),
  `Imbalance strategy` = c("Class weights + SMOTE", "Class weights",
                            "Class weights", "scale_pos_weight",
                            "Class weights + SMOTE", "Class prior")
)
kable(model_tbl, booktabs = TRUE) %>%
  kable_styling(full_width = TRUE, font_size = 14) %>%
  column_spec(1, bold = TRUE, width = "6em") %>%
  column_spec(3, width = "18em") %>%
  row_spec(0, bold = TRUE, background = "#2E74B5", color = "white")
```

*All models trained via `caret` with 5-fold CV. Primary metric: **Macro-F1**. Hard constraint: **Fatal Recall ≥ 0.50**.*

## Model Fitting: Logistic Regression

<small style="color:#2E74B5">**Rubric criterion:** Model Fitting  \|  2 pts</small>

```{r results-lr, fig.height=3.5}
# TODO: [member name] — confusion matrix heatmap + coefficient plot or ROC curves
# Train on: df_train | CV: cv_ctrl_smote | Evaluate on: df_test
```

## Model Fitting: Decision Tree

<small style="color:#2E74B5">**Rubric criterion:** Model Fitting  \|  2 pts</small>

```{r results-dt, fig.height=3.5}
# TODO: [member name] — rpart.plot tree diagram + confusion matrix heatmap
# Train on: df_train | CV: cv_ctrl | Evaluate on: df_test
```

## Model Fitting: Random Forest

<small style="color:#2E74B5">**Rubric criterion:** Model Fitting  \|  2 pts</small>

```{r results-rf, fig.height=3.5}
# TODO: [member name] — variable importance plot (top 15) + confusion matrix heatmap
# Train on: df_train | CV: cv_ctrl | Evaluate on: df_test
```

## Model Fitting: XGBoost

<small style="color:#2E74B5">**Rubric criterion:** Model Fitting  \|  2 pts</small>

```{r results-xgb, fig.height=3.5}
# TODO: [member name] — variable importance (by Gain) + training loss curve + confusion matrix
# Train on: df_train | CV: cv_ctrl | Evaluate on: df_test
```

## Model Fitting: KNN & Naive Bayes

<small style="color:#2E74B5">**Rubric criterion:** Model Fitting  \|  2 pts</small>

```{r results-knn-nb, fig.height=3.5}
# TODO: [member name] — k vs Macro-F1 curve (KNN) + side-by-side confusion matrices
# KNN: cv_ctrl_smote | NB: cv_ctrl_smote | Evaluate both on: df_test
```

## Model Improvement: Hyperparameter Tuning

<small style="color:#2E74B5">**Rubric criterion:** Model Improvement  \|  2 pts</small>

```{r slide-improvement}
# TODO: [member name] — tuning table (model, parameter, range, best value) + before/after metric comparison
```

## Model Evaluation

<small style="color:#2E74B5">**Rubric criterion:** Model Evaluation  \|  2 pts</small>

```{r slide-evaluation}
# TODO: [member name] — comparison table (6 models x Macro-F1, Fatal Recall, Weighted F1, Kappa)
```

## Model Selection

<small style="color:#2E74B5">**Rubric criterion:** Model Selection  \|  2 pts</small>

```{r slide-selection}
# TODO: [member name] — best model name, justification, top predictors from variable importance
```

## Final Results & Conclusion

<small style="color:#2E74B5">**Rubric criterion:** Final Results & Conclusion  \|  2 pts</small>

```{r slide-conclusion}
# TODO: [member name] — answer to RQ, key findings, practical implications for stakeholders
```

## Reflection & Future Directions

<small style="color:#2E74B5">**Rubric criterion:** Final Results & Conclusion  \|  2 pts (continued)</small>

```{r slide-reflection}
# TODO: [member name] — limitations, future research directions
```

## References

```{r slide-references}
# TODO: [member name] — APA 7th references for all sources cited across slides
# AI acknowledgement to be added as a <small> footnote at the bottom
```
