---
title: ""
author: ""
output:
  html_document:
    code_folding: hide
---

```{r library, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(readr)
library(forcats)
library(scales)
library(knitr)
library(caret)
library(nnet)
library(rpart)
library(rpart.plot)
library(randomForest)
library(xgboost)
library(kknn)
library(e1071)
library(themis)
library(pROC)
library(flextable)
```

```{css, echo=FALSE}
body { font-family: "Times New Roman", Times, serif; font-size: 10pt; line-height: 1.3; }
h2 { margin-top: 0.4em; margin-bottom: 0.15em; }
h3 { margin-top: 0.3em; margin-bottom: 0.1em; }
h4 { margin-top: 0.25em; margin-bottom: 0.1em; }
p  { margin-top: 0.2em; margin-bottom: 0.2em; }
h1.title { display: none; }
h4.author { display: none; }
h4.date   { display: none; }
```

```{r load-data, message=FALSE, warning=FALSE, results='hide'}
df <- read.csv("nsw_road_crash_merged_2020_2024.csv", check.names = FALSE)
cat("Dimensions:", nrow(df), "rows x", ncol(df), "cols\n")
str(df)
```

---

### 1 Problem Framing & Research Question

<!-- TODO: [member name] — RQ statement, motivation, 3 stakeholder groups (~0.5 page) -->

---

### 2 Dataset & Justification

<!-- TODO: [member name] — data source, merge process, 4-requirement justification table (~0.5 page) -->

---

### 3 EDA & Insights

<!-- TODO: [member name] — 4 ggplot2 plots with key finding commentary per plot (~1 page) -->

---

### 4 Data Preprocessing

All data-dependent steps are performed **within cross-validation folds** to prevent data leakage (James et al., 2021).

##### Step 1 — Data Integration & TU-Level Feature Aggregation

```{r step1-integration, message=FALSE, warning=FALSE, results='hide'}
# Each row is one traffic unit — summarise binary flags per crash, then join back
tu_flags <- df %>%
  group_by(`Crash ID`) %>%
  summarise(
    has_motorcycle    = as.integer(any(`TU type group` == "Motorcycle")),
    has_heavy_vehicle = as.integer(any(`TU type group` %in% c("Articulated truck", "Heavy rigid truck", "Bus"))),
    has_pedestrian    = as.integer(any(`TU type group` == "Pedestrian")),
    has_cyclist       = as.integer(any(`TU type group` == "Pedal cycle")),
    tu_count          = n(),
    .groups = "drop"
  )

df <- df %>% left_join(tu_flags, by = "Crash ID")
cat("TU flags added. Rows:", nrow(df), "| New cols:", ncol(df), "\n")
```

##### Step 2 — Feature Removal

```{r step2-removal, message=FALSE, warning=FALSE, results='hide'}
# Leakage: post-crash outcome counts that directly encode the target
leakage_cols <- c("No. killed", "No. seriously injured",
                  "No. moderately injured", "No. minor-other injured")

# Redundant: detailed severity is a finer version of the target column
redundant_target_cols <- c("Degree of crash - detailed")

# Identifiers and free-text location fields (no predictive signal)
id_cols <- c("Crash ID", "Traffic unit ID", "Street of crash",
             "Street of travel", "Town", "Identifying feature",
             "Identifying feature type", "Conurbation 1")

# High-NA columns (>75% missing — confirmed from inspection)
high_na_cols <- c("Primary hazardous feature", "Primary temporary feature",
                  "Object hit 2", "DCA supplement", "Object hit 1",
                  "Primary permanent feature")

redundant_cols <- c("Route no.", "RUM - code", "RUM - description",
                    "DCA - code", "DCA - description")

drop_cols <- unique(c(leakage_cols, redundant_target_cols, id_cols, high_na_cols, redundant_cols))
df_clean  <- df %>% select(-any_of(drop_cols))
cat("Columns after removal:", ncol(df_clean), "\n")
```

##### Step 3 — Missing Value Imputation

```{r step3-imputation, message=FALSE, warning=FALSE, results='hide'}
df_clean <- df_clean %>%
  mutate(
    `Other TU type` = ifelse(is.na(`Other TU type`), "Unknown", `Other TU type`),
    across(where(is.character), ~ replace_na(., "Unknown")),
    across(where(is.numeric),   ~ replace_na(., median(., na.rm = TRUE)))
  )
cat("Missing values after imputation:", sum(is.na(df_clean)), "\n")
```

##### Step 4 — Feature Engineering

```{r step4-engineering, message=FALSE, warning=FALSE, results='hide'}
df_clean <- df_clean %>%
  mutate(
    crash_hour   = as.integer(sub(":.*", "", `Two-hour intervals`)),
    is_weekend   = as.integer(`Day of week of crash` %in% c("Saturday", "Sunday")),
    is_rush_hour = as.integer(crash_hour %in% c(7, 8, 9, 16, 17, 18)),
    is_night     = as.integer(`Natural lighting` %in% c("Darkness", "Dawn", "Dusk")),
    is_wet       = as.integer(`Surface condition` %in% c("Wet", "Snow or ice")),
    speed_band   = case_when(
      `Speed limit` %in% c("10 km/h","20 km/h","30 km/h","40 km/h","50 km/h") ~ "low_50",
      `Speed limit` %in% c("60 km/h","70 km/h","80 km/h")                     ~ "mid_60_80",
      `Speed limit` %in% c("90 km/h","100 km/h")                              ~ "high_90_100",
      `Speed limit` == "110 km/h"                                              ~ "very_high_110",
      TRUE                                                                     ~ "unknown"
    )
  )
cat("Engineered features added. Columns:", ncol(df_clean), "\n")
```

##### Step 5 — Outlier Detection

```{r step5-outliers, message=FALSE, warning=FALSE, results='hide'}
p99_dist <- quantile(df_clean$Distance, 0.99, na.rm = TRUE)
df_clean <- df_clean %>%
  mutate(Distance = pmin(Distance, p99_dist)) %>%
  filter(is.na(Latitude)  | (Latitude  >= -37.5 & Latitude  <= -28.2),
         is.na(Longitude) | (Longitude >= 141.0  & Longitude <= 153.6))
cat("Rows after outlier removal:", nrow(df_clean), "\n")
```

##### Step 6 — Encoding & LGA Pooling

```{r step6-encoding, message=FALSE, warning=FALSE, results='hide'}
lga_freq  <- prop.table(table(df_clean$LGA))
rare_lgas <- names(lga_freq[lga_freq < 0.01])
df_clean  <- df_clean %>%
  mutate(LGA = ifelse(LGA %in% rare_lgas, "Other", LGA))

df_model <- df_clean %>%
  mutate(across(where(is.character), as.factor)) %>%
  mutate(`Degree of crash` = factor(`Degree of crash`,
           levels = c("Fatal", "Injury", "Non-casualty (towaway)")))

cat("Class distribution:\n")
print(prop.table(table(df_model$`Degree of crash`)))
cat("LGA levels after pooling:", nlevels(df_model$LGA), "\n")
```

##### Step 7 — Train/Test Split & Temporal Holdout

```{r step7-split, message=FALSE, warning=FALSE, results='hide'}
# 2024 sealed as temporal holdout — evaluated once at the very end only
df_holdout_2024  <- df_model %>% filter(`Reporting year` == 2024)
df_model_2020_23 <- df_model %>% filter(`Reporting year` != 2024)

set.seed(42)
train_idx <- createDataPartition(df_model_2020_23$`Degree of crash`, p = 0.8, list = FALSE)
df_train  <- df_model_2020_23[ train_idx, ]
df_test   <- df_model_2020_23[-train_idx, ]

cat("Train:", nrow(df_train), "| Test:", nrow(df_test), "| Holdout:", nrow(df_holdout_2024), "\n")
print(prop.table(table(df_train$`Degree of crash`)))
```

##### Step 8 — Class Imbalance Handling

```{r step8-smote, message=FALSE, warning=FALSE, results='hide'}
# Inverse frequency weights for models supporting class weights (LR, DT, RF)
class_weights <- 1 / prop.table(table(df_train$`Degree of crash`))
class_weights <- class_weights / sum(class_weights)
cat("Class weights:\n"); print(round(class_weights, 4))
# SMOTE applied inside CV fold for KNN and Naive Bayes via cv_ctrl_smote
```

##### Step 9 — Cross-Validation Setup

```{r step9-cv, message=FALSE, warning=FALSE, results='hide'}
cv_ctrl <- trainControl(
  method = "cv", number = 5, classProbs = TRUE,
  summaryFunction = multiClassSummary, savePredictions = "final"
)
cv_ctrl_smote <- trainControl(
  method = "cv", number = 5, classProbs = TRUE,
  summaryFunction = multiClassSummary, savePredictions = "final",
  sampling = "smote"
)
cat("CV controls ready.\n")
```

---

### 5 Model Fitting

Six models selected to cover a range of complexity and interpretability. All implemented in R via `caret` with 5-fold CV. Primary metric: **Macro-F1**. Hard constraint: **Fatal Recall ≥ 0.50**.

#### 5.1 Multinomial Logistic Regression

```{r model-lr, message=FALSE, warning=FALSE}
# Encoding & scaling: add preProcess = "nzv" in train()
# method = "multinom" handles factors via dummy variables internally
# Train on: df_train | CV control: cv_ctrl_smote | Weights: class_weights
# Evaluate on: df_test | Temporal check: df_holdout_2024

```

#### 5.2 Decision Tree

```{r model-dt, message=FALSE, warning=FALSE}
# Encoding & scaling: none required — rpart handles R factors natively
# Train on: df_train | CV control: cv_ctrl | Weights: class_weights
# Evaluate on: df_test | Temporal check: df_holdout_2024

```

#### 5.3 Random Forest

```{r model-rf, message=FALSE, warning=FALSE}
# Encoding & scaling: none required — randomForest handles R factors natively
# Train on: df_train | CV control: cv_ctrl | Weights: classwt = class_weights
# Evaluate on: df_test | Temporal check: df_holdout_2024

```

#### 5.4 XGBoost

```{r model-xgb, message=FALSE, warning=FALSE}
# Encoding: requires numeric matrix — use model.matrix(~ . - 1, data = df_train) to one-hot encode
# Train on: df_train | CV control: cv_ctrl | Weights: scale_pos_weight
# Evaluate on: df_test | Temporal check: df_holdout_2024

```

#### 5.5 K-Nearest Neighbours

```{r model-knn, message=FALSE, warning=FALSE}
# Encoding & scaling: add preProcess = c("center", "scale") in train()
# Train on: df_train | CV control: cv_ctrl_smote | Weights: via SMOTE
# Evaluate on: df_test | Temporal check: df_holdout_2024

```

#### 5.6 Naive Bayes

```{r model-nb, message=FALSE, warning=FALSE}
# Encoding & scaling: none required — naive_bayes handles factors directly
# Train on: df_train | CV control: cv_ctrl_smote | Weights: via SMOTE
# Evaluate on: df_test | Temporal check: df_holdout_2024

```

---

### 6 Model Improvement

<!-- TODO: [member name] — per-model tuning (§6.1–§6.6): tuning grid, CV curve, best params, before/after comparison -->

---

### 7 Model Evaluation & Selection

<!-- TODO: [member name] — comparison table (6 models × 4 metrics), best model selection + justification -->

---

### 8 Final Results & Conclusion

<!-- TODO: [member name] — answer RQ, top predictors, practical implications, process reflection -->

---

### References

<!-- TODO: [member name] — APA 7th format references for all sources cited -->

### Acknowledgement of AI

<!-- TODO: [member name] — AI acknowledgement statement -->
