---
title: "NSW Road Crash Severity Prediction"
subtitle: "STAT5003 — Group 07, Workshop 12"
author: "Group 07, Workshop 12"
date: "May 2026"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.align = "center", out.width = "90%")

library(ggplot2)
library(dplyr)
library(readr)
library(forcats)
library(scales)
library(knitr)
library(kableExtra)
library(tidyr)
```

```{r load-data, include=FALSE}
df_raw <- read_csv("nsw_road_crash_merged_2020_2024.csv", show_col_types = FALSE)

df_raw <- df_raw %>%
  rename(severity = `Degree of crash`) %>%
  mutate(severity = factor(severity,
    levels = c("Fatal", "Injury", "Non-casualty (towaway)"),
    labels = c("Fatal", "Injury", "Non-casualty")))
```

## Problem Framing & Research Question

<small style="color:#2E74B5">**Rubric criterion:** Problem Framing & Research Question &nbsp;|&nbsp; 2 pts</small>

```{r slide-rq}
# TODO: [member name] — RQ statement, motivation, stakeholder value
```

## EDA & Data Understanding: Dataset Overview

<small style="color:#2E74B5">**Rubric criterion:** EDA & Data Understanding &nbsp;|&nbsp; 2 pts</small>

```{r slide-dataset}
# TODO: [member name] — dataset summary table (source, records, variables, target, date range)
```

## EDA & Data Understanding: Class Imbalance

<small style="color:#2E74B5">**Rubric criterion:** EDA & Data Understanding &nbsp;|&nbsp; 2 pts</small>

```{r slide-imbalance, fig.height=3.5}
# TODO: [member name] — bar chart of Fatal / Injury / Non-casualty proportions
```

## EDA & Data Understanding: Speed vs Severity

<small style="color:#2E74B5">**Rubric criterion:** EDA & Data Understanding &nbsp;|&nbsp; 2 pts</small>

```{r slide-speed, fig.height=3.5}
# TODO: [member name] — stacked bar chart of severity proportion by speed limit band
```

## EDA & Data Understanding: High-Risk Scenarios

<small style="color:#2E74B5">**Rubric criterion:** EDA & Data Understanding &nbsp;|&nbsp; 2 pts</small>

```{r slide-highrisk, fig.height=3.5}
# TODO: [member name] — fatal rate by speed band and lighting condition
```

## EDA & Data Understanding: Temporal Patterns

<small style="color:#2E74B5">**Rubric criterion:** EDA & Data Understanding &nbsp;|&nbsp; 2 pts</small>

```{r slide-temporal, fig.height=3.5}
# TODO: [member name] — crash frequency heatmap by day of week and time interval
```

## Data Preprocessing: Pipeline

<small style="color:#2E74B5">**Rubric criterion:** Data Preprocessing &nbsp;|&nbsp; 2 pts</small>

**Steps 1–5: Cleaning & Feature Engineering**

1. **TU Integration** — aggregated traffic-unit flags per crash (`has_motorcycle`, `has_heavy_vehicle`, `has_pedestrian`, `has_cyclist`) using `group_by + summarise + left_join`
2. **Feature removal** — dropped leakage columns (post-crash injury counts), identifiers, and variables with >75% missing values; dropped `Degree of crash - detailed` (redundant with target)
3. **Imputation** — removed rows >50% missing; categorical → `"Unknown"`; numeric → median
4. **Feature engineering** — `is_weekend`, `is_rush_hour`, `is_night`, `is_wet`, speed bands — all motivated by EDA findings
5. **Outlier handling** — NSW geographic bounds filter; winsorise `Distance` at 99th percentile

**Steps 6–9: Modelling Setup**

6. **Encoding** — trees use native R `factor`; KNN/LR use one-hot; rare LGA levels → `"Other"`
7. **Scaling** — Z-score standardisation within fold for KNN; trees are scale-invariant
8. **Class imbalance** — inverse-frequency class weights (all models); SMOTE applied **inside training fold only**
9. **Split** — 2024 data sealed as temporal holdout; 80/20 stratified split on 2020–2023; 5-fold CV

## Data Preprocessing: Feature Engineering & Class Balancing

<small style="color:#2E74B5">**Rubric criterion:** Data Preprocessing &nbsp;|&nbsp; 2 pts</small>

**Engineered features — all motivated by EDA**

| Feature | Derivation | EDA motivation |
|---|---|---|
| `speed_band` | Cut `Speed limit` into ≤50 / 60–80 / 90–100 / 110 | Fatal rate rises sharply above 100 km/h |
| `is_night` | `Natural lighting` ∈ {Darkness, Dawn, Dusk} | Dark conditions double fatal risk |
| `is_rush_hour` | `Two-hour intervals` ∈ {7–9, 16–18} | Commute peaks dominate crash volume |
| `is_weekend` | `Day of week of crash` ∈ Saturday, Sunday | Weekend crashes differ in severity profile |
| `is_wet` | `Surface condition` contains "Wet" | Wet roads linked to skidding fatalities |
| `has_motorcycle` | TU-level aggregated flag | Motorcyclists over-represented in fatals |
| `has_pedestrian` | TU-level aggregated flag | Pedestrian crashes nearly always severe |

**Handling class imbalance**

- Raw Fatal share ~1.4% → accuracy-maximising models ignore Fatal entirely
- Strategy: **inverse-frequency class weights** (all models) + **SMOTE** inside training fold (LR, KNN)
- SMOTE raises synthetic Fatal share to ~10% in training only — test set remains unmodified

## Model Fitting: Approach & Setup

<small style="color:#2E74B5">**Rubric criterion:** Model Fitting &nbsp;|&nbsp; 2 pts</small>

```{r model-overview}
model_tbl <- tibble::tibble(
  Model              = c("Multinomial Logistic Regression", "Decision Tree",
                         "Random Forest", "XGBoost",
                         "K-Nearest Neighbours", "Naive Bayes"),
  `R package`        = c("nnet", "rpart", "randomForest", "xgboost", "kknn", "e1071"),
  `Why chosen`       = c(
    "Interpretable baseline; log-odds show predictor direction",
    "Fully interpretable decision rules; native categorical handling",
    "Ensemble; robust to noise; provides variable importance",
    "Gradient boosting; typically strongest on tabular data",
    "Non-parametric; captures local structure without assumptions",
    "Probabilistic baseline; reveals feature independence violations"
  ),
  `Imbalance strategy` = c("Class weights + SMOTE", "Class weights",
                            "Class weights", "scale_pos_weight",
                            "Class weights + SMOTE", "Class prior")
)
kable(model_tbl, booktabs = TRUE) %>%
  kable_styling(full_width = TRUE, font_size = 14) %>%
  column_spec(1, bold = TRUE, width = "6em") %>%
  column_spec(3, width = "18em") %>%
  row_spec(0, bold = TRUE, background = "#2E74B5", color = "white")
```

*All models trained via `caret` with 5-fold CV. Primary metric: **Macro-F1**. Hard constraint: **Fatal Recall ≥ 0.50**.*

## Model Fitting: Logistic Regression

<small style="color:#2E74B5">**Rubric criterion:** Model Fitting &nbsp;|&nbsp; 2 pts</small>

```{r results-lr, fig.height=3.5}
# TODO: [member name] — confusion matrix heatmap + coefficient plot or ROC curves
# Train on: df_train | CV: cv_ctrl_smote | Evaluate on: df_test
```

## Model Fitting: Decision Tree

<small style="color:#2E74B5">**Rubric criterion:** Model Fitting &nbsp;|&nbsp; 2 pts</small>

```{r results-dt, fig.height=3.5}
# TODO: [member name] — rpart.plot tree diagram + confusion matrix heatmap
# Train on: df_train | CV: cv_ctrl | Evaluate on: df_test
```

## Model Fitting: Random Forest

<small style="color:#2E74B5">**Rubric criterion:** Model Fitting &nbsp;|&nbsp; 2 pts</small>

```{r results-rf, fig.height=3.5}
# TODO: [member name] — variable importance plot (top 15) + confusion matrix heatmap
# Train on: df_train | CV: cv_ctrl | Evaluate on: df_test
```

## Model Fitting: XGBoost

<small style="color:#2E74B5">**Rubric criterion:** Model Fitting &nbsp;|&nbsp; 2 pts</small>

```{r results-xgb, fig.height=3.5}
# TODO: [member name] — variable importance (by Gain) + training loss curve + confusion matrix
# Train on: df_train | CV: cv_ctrl | Evaluate on: df_test
```

## Model Fitting: KNN & Naive Bayes

<small style="color:#2E74B5">**Rubric criterion:** Model Fitting &nbsp;|&nbsp; 2 pts</small>

```{r results-knn-nb, fig.height=3.5}
# TODO: [member name] — k vs Macro-F1 curve (KNN) + side-by-side confusion matrices
# KNN: cv_ctrl_smote | NB: cv_ctrl_smote | Evaluate both on: df_test
```

## Model Improvement: Hyperparameter Tuning

<small style="color:#2E74B5">**Rubric criterion:** Model Improvement &nbsp;|&nbsp; 2 pts</small>

```{r slide-improvement}
# TODO: [member name] — tuning table (model, parameter, range, best value) + before/after metric comparison
```

## Model Evaluation

<small style="color:#2E74B5">**Rubric criterion:** Model Evaluation &nbsp;|&nbsp; 2 pts</small>

```{r slide-evaluation}
# TODO: [member name] — comparison table (6 models x Macro-F1, Fatal Recall, Weighted F1, Kappa)
```

## Model Selection

<small style="color:#2E74B5">**Rubric criterion:** Model Selection &nbsp;|&nbsp; 2 pts</small>

```{r slide-selection}
# TODO: [member name] — best model name, justification, top predictors from variable importance
```

## Final Results & Conclusion

<small style="color:#2E74B5">**Rubric criterion:** Final Results & Conclusion &nbsp;|&nbsp; 2 pts</small>

```{r slide-conclusion}
# TODO: [member name] — answer to RQ, key findings, practical implications for stakeholders
```

## Reflection & Future Directions

<small style="color:#2E74B5">**Rubric criterion:** Final Results & Conclusion &nbsp;|&nbsp; 2 pts (continued)</small>

```{r slide-reflection}
# TODO: [member name] — limitations, future research directions
```

## References

```{r slide-references}
# TODO: [member name] — APA 7th references for all sources cited across slides
# AI acknowledgement to be added as a <small> footnote at the bottom
```
