---
title: "NSW Road Crash Severity Prediction"
subtitle: "STAT5003 — Group 07, Workshop 12"
author: "Group 07, Workshop 12."
date: "May 2026"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.align = "center", out.width = "90%")

library(ggplot2)
library(dplyr)
library(readr)
library(forcats)
library(scales)
library(knitr)
library(caret)
library(rpart.plot)
library(randomForest)
library(pROC)
```

## Slide 1 — Research Question

**Can we predict NSW road crash severity given environmental and vehicle conditions?**

- **Target:** Degree of crash — *Fatal / Injury / Non-casualty* (3-class)
- **Predictors:** Weather, lighting, road surface, speed limit, TU type, manoeuvre
- **Data:** 170,962 crash records, 2020–2024 (Transport for NSW)

**Why it matters:**

- NSW: 340 road fatalities in 2024, >⅔ in regional areas
- Traffic Authorities → targeted interventions
- Emergency Responders → evidence-based resource allocation
- Policymakers & Insurers → precise policy and pricing

## Slide 2 — Dataset Overview

```{r dataset-overview}

```

## Slide 3 — Class Imbalance

```{r class-imbalance, fig.height=3.5}

```

*Fatal class (~1.4%) requires imbalance-aware modelling — accuracy alone is misleading.*

## Slide 4 — Key EDA: Speed vs Severity

```{r eda-speed, fig.height=3.5}

```

*100–110 km/h zones show the highest Fatal proportions — strongest single predictor.*

## Slide 5 — Key EDA: High-Risk Scenarios

```{r eda-highrisk}

```

*High-speed + darkness + non-urban = highest fatal risk. Justifies interaction-based tree models.*

## Slide 6 — Key EDA: Temporal Patterns

```{r eda-temporal, fig.height=3.5}

```

*Peaks during commute hours. Severity proportions stable across years — imbalance is structural.*

## Slide 7 — Preprocessing Pipeline

**Steps 1–5**

1. **Integration** — TU flags per crash (`has_motorcycle`, `has_heavy_vehicle`)
2. **Feature removal** — leakage cols, identifiers, >75% NA
3. **Imputation** — drop >50% NA; categorical → "Unknown"; numeric → median
4. **Engineering** — `is_weekend`, `is_rush_hour`, `is_night`, `is_wet`, speed bands
5. **Outliers** — NSW geo bounds; winsorise Distance at 99th percentile

**Steps 6–9**

6. **Encoding** — trees: R `factor`; KNN/LR: one-hot; rare LGA → "Other"
7. **Scaling** — Z-score within fold for KNN; trees scale-invariant
8. **SMOTE** — applied inside training fold only (Fatal minority)
9. **Split** — 80/20 stratified; nested 5-fold CV (inner: tune, outer: estimate)

## Slide 8 — Feature Engineering

```{r feature-eng-table}

```

## Slide 9 — Handling Class Imbalance (SMOTE)

```{r smote-plot, fig.height=3.2}

```

*SMOTE applied **inside** training fold only to prevent leakage. Fatal class upsampled to ~10%.*

## Slide 10 — Model Overview

```{r model-overview}

```

*All models trained via `caret` with 5-fold CV. Primary metric: **Macro-F1**. Hard constraint: **Fatal Recall ≥ 0.50**.*

## Slide 11 — Logistic Regression Results

```{r results-lr, fig.height=3.5}

```

## Slide 12 — Decision Tree Results

```{r results-dt, fig.height=3.5}

```

## Slide 13 — Random Forest Results

```{r results-rf, fig.height=3.5}

```

## Slide 14 — XGBoost Results

```{r results-xgb, fig.height=3.5}

```

## Slide 15 — KNN & Naive Bayes Results

```{r results-knn-nb, fig.height=3.5}

```

## Slide 16 — Model Comparison

```{r model-comparison}

```

*Bold = best model. Models failing Fatal Recall ≥ 0.50 are deprioritised regardless of overall F1.*

## Slide 17 — Best Model & Answer to RQ

**Selected model:**

**Answer to RQ:** Given environmental conditions and traffic unit characteristics, crash severity **can / cannot** be reliably predicted. Key findings:

- **Top predictors:** [from RF/XGBoost variable importance]
- **Fatal recall:** Best model achieves ≥ 0.50 — meets operational constraint
- **Interaction effects:** High-speed + rural + darkness combination drives fatal risk
- **TU-level flags:** `has_motorcycle` and `has_pedestrian` add predictive value

## Slide 18 — Conclusion

- Trained 6 classifiers on 170,962 NSW crash records (2020–2024)
- Addressed severe class imbalance (Fatal ~1.4%) via SMOTE + class weights
- **Best model:** [Model name] — Macro-F1 = ?, Fatal Recall = ?
- Speed limit and urbanisation are the strongest predictors of crash severity
- High-speed, non-urban, low-light conditions are the highest-risk profile

**Practical value:**

- Traffic Authorities: prioritise enforcement in 100–110 km/h rural corridors
- Emergency Responders: pre-position resources based on predicted fatal risk

## Slide 19 — Reflection & Scope for Future Research

**Limitations**

- Fatal class severely underrepresented even after SMOTE
- No spatial/network features (road geometry, curve radius)
- Cross-sectional data — no causal inference
- LGA pooling loses granular spatial signal
- KNN and LR struggle with high-cardinality categoricals

**Future Directions**

- Deep learning (TabNet, MLP) for richer feature interactions
- Spatial modelling — GWR or spatial random effects by LGA
- Real-time crash risk prediction integrated with TfNSW traffic data
- Temporal models to capture trend and seasonality
- Explainability via SHAP values for stakeholder communication

## References

Chawla, N. V., Bowyer, K. W., Hall, L. O., & Kegelmeyer, W. P. (2002). SMOTE: Synthetic minority over-sampling technique. *JAIR*, 16, 321–357.

James, G., Witten, D., Hastie, T., & Tibshirani, R. (2021). *An introduction to statistical learning: With applications in R* (2nd ed.). Springer.

Transport for NSW. (2024). *NSW road crash data 2020–2024* [Dataset]. https://www.data.gov.au/data/dataset/nsw-2-nsw-crash-data

Transport for NSW. (2025). *NSW road safety progress report 2024*. Centre for Road Safety.

---

**Acknowledgement of AI:** Group 07 (Workshop 12) declares all ideas, analysis, and written content are our own. We used ChatGPT (GPT-5.3) and Claude Sonnet 4.6 for code refinement, grammar checking, and rubric review only.
