---
title: "NSW Road Crash Severity Prediction"
subtitle: "STAT5003  -  Group 07, Workshop 12"
author: "Sanskar | Robert | Haipeng | Chen | Yihan"
date: "May 2026"
output:
  html_document:
    code_folding: hide
---


```{r library, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(readr)
library(forcats)
library(scales)
library(knitr)
library(caret)
library(nnet)
library(rpart)
library(rpart.plot)
library(randomForest)
library(xgboost)
library(kknn)
library(e1071)
library(themis)
library(pROC)
library(flextable)
library(MLmetrics)

# Create plots output folder  -  all ggsave() calls write here for use in Presentation Rmd
dir.create("plots", showWarnings = FALSE)
```

```{css, echo=FALSE}
body { font-family: "Times New Roman", Times, serif; font-size: 10pt; line-height: 1.3; }
h2 { margin-top: 0.4em; margin-bottom: 0.15em; }
h3 { margin-top: 0.3em; margin-bottom: 0.1em; }
h4 { margin-top: 0.25em; margin-bottom: 0.1em; }
p  { margin-top: 0.2em; margin-bottom: 0.2em; }
h1.title { display: none; }
h4.author { display: none; }
h4.date   { display: none; }
```

```{r load-data, message=FALSE, warning=FALSE, results='hide'}
df <- read.csv("nsw_road_crash_merged_2020_2024.csv", check.names = FALSE)
cat("Dimensions:", nrow(df), "rows x", ncol(df), "cols\n")
str(df)
```

------------------------------------------------------------------------

### 1 Problem Framing & Research Question

#### 1.1 Research Question
Given environmental conditions (weather, natural lighting, road surface, speed limit) and traffic unit characteristics (type of traffic unit, manoeuvre, number of units involved), can we predict the severity of a road crash in NSW as a three-class classification problem: Fatal, Injury, or Non-casualty?

#### 1.2 Motivation
NSW recorded 340 road fatalities in 2024, with over two-thirds occurring in regional areas (Transport for NSW, 2025), highlighting the need for data-driven approaches to understanding crash severity. Fatal crashes represent only approximately 1.4% of all crashes, introducing severe class imbalance that must be carefully addressed during modelling and evaluation.

A predictive model for crash severity has direct real-world value for the following stakeholders:
NSW Traffic Management Authorities: By identifying key combinations of environmental and traffic characteristics associated with severe crashes, authorities can make targeted decisions such as adjusting speed limits and improving road design in high-risk areas.

Emergency Response Teams: Predicting crash severity based on environmental conditions in the early stages of an incident can support faster and more evidence-based allocation of ambulance and emergency resources.
Policymakers and Insurance Institutions: Understanding which traffic unit types and environmental conditions are associated with higher severity risk can inform more precise road safety policies and insurance pricing models.

------------------------------------------------------------------------

### 2 Dataset & Justification

#### 2.1 Dataset Overview

```{r dataset-overview, echo=FALSE, message=FALSE, warning=FALSE}
overview <- data.frame(
  Item = c(
    "Time Period",
    "Data Sources",
    "Merge Key",
    "Total Records",
    "Total Variables",
    "Target Variable",
    "Fatal",
    "Injury",
    "Non-casualty"
  ),
  Detail = c(
    "2020 - 2024",
    "CRASH.xlsx + TRAFFIC UNIT.xlsx (TfNSW via data.gov.au)",
    "Crash ID (left join)",
    format(nrow(df), big.mark = ","),
    as.character(ncol(df)),
    "Degree of Crash (3-class classification)",
    paste0(format(sum(df$`Degree of crash` == "Fatal"), big.mark = ","),
           " (", round(mean(df$`Degree of crash` == "Fatal") * 100, 1), "%)"),
    paste0(format(sum(df$`Degree of crash` == "Injury"), big.mark = ","),
           " (", round(mean(df$`Degree of crash` == "Injury") * 100, 1), "%)"),
    paste0(format(sum(df$`Degree of crash` == "Non-casualty (towaway)"), big.mark = ","),
           " (", round(mean(df$`Degree of crash` == "Non-casualty (towaway)") * 100, 1), "%)")
  )
)

knitr::kable(overview, col.names = c("Item", "Detail"),
             caption = "Table 1: Dataset Overview")
```

------------------------------------------------------------------------

### 3 EDA & Insights
```{r 311-selected, fig.width=12, fig.height=4, message=FALSE, warning=FALSE}
library(gridExtra)

p1 <- ggplot(df, aes(x = fct_infreq(`Degree of crash`))) +
    geom_bar() +
    geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.5, size = 3) +
    labs(title = "Distribution of Crash Severity", x = "Severity", y = "Count") +
    theme_minimal() + theme(legend.position = "none")

p2 <- ggplot(df, aes(x = `Speed limit`)) +
  geom_bar(fill = "tomato") +
  labs(title = "Distribution of Speed Limit", x = "Speed Limit", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p3 <- ggplot(df, aes(x = fct_infreq(`TU type group`))) +
    geom_bar(fill = "steelblue") +
    labs(title = "Distribution of TU Type", x = "TU Type", y = "Count") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

p4 <- ggplot(df, aes(x = fct_infreq(Urbanisation))) +
  geom_bar(fill = "purple") +
  labs(title = "Distribution of Urbanisation", x = "Urbanisation", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

grid.arrange(p1, p2, p3, p4, nrow = 1)
```

- **Crash Severity:** Fatal crashes represent only 1.4% of records, indicating severe class imbalance that must be addressed during modelling.
- **Speed Limit:** Crashes are heavily concentrated in 50-60 km/h urban zones, reflecting higher traffic volume rather than elevated per-trip risk.
- **Traffic Unit Type:** Car/car derivatives dominate (~114,477), with pedestrians and motorcycles being rare but disproportionately associated with severe outcomes.
- **Urbanisation:** Sydney metro accounts for the most crashes (98,975), while Country non-urban regions show higher fatal proportions despite lower overall volume.

------------------------------------------------------------------------

### 4 Data Preprocessing

All data-dependent steps are performed **within cross-validation folds** to prevent data leakage (James et al., 2021).

#### 4.1 Data Integration & TU-Level Feature Aggregation

```{r step1-integration, message=FALSE, warning=FALSE, results='hide'}
# Each row is one traffic unit  -  summarise binary flags per crash, then join back
tu_flags <- df %>%
  group_by(`Crash ID`) %>%
  summarise(
    has_motorcycle    = as.integer(any(`TU type group` == "Motorcycle")),
    has_heavy_vehicle = as.integer(any(`TU type group` %in% c("Articulated truck", "Heavy rigid truck", "Bus"))),
    has_pedestrian    = as.integer(any(`TU type group` == "Pedestrian")),
    has_cyclist       = as.integer(any(`TU type group` == "Pedal cycle")),
    tu_count          = n(),
    .groups = "drop"
  )

df <- df %>% left_join(tu_flags, by = "Crash ID")
cat("TU flags added. Rows:", nrow(df), "| New cols:", ncol(df), "\n")
```

#### 4.2 Feature Removal

```{r step2-removal, message=FALSE, warning=FALSE, results='hide'}
# Leakage: post-crash outcome counts that directly encode the target
leakage_cols <- c("No. killed", "No. seriously injured",
                  "No. moderately injured", "No. minor-other injured")

# Redundant: detailed severity is a finer version of the target column
redundant_target_cols <- c("Degree of crash - detailed")

# Identifiers and free-text location fields (no predictive signal)
id_cols <- c("Crash ID", "Traffic unit ID", "Street of crash",
             "Street of travel", "Town", "Identifying feature",
             "Identifying feature type", "Conurbation 1")

# High-NA columns (>75% missing  -  confirmed from inspection)
high_na_cols <- c("Primary hazardous feature", "Primary temporary feature",
                  "Object hit 2", "DCA supplement", "Object hit 1",
                  "Primary permanent feature")

redundant_cols <- c("Route no.", "RUM - code", "RUM - description",
                    "DCA - code", "DCA - description")

drop_cols <- unique(c(leakage_cols, redundant_target_cols, id_cols, high_na_cols, redundant_cols))
df_clean  <- df %>% select(-any_of(drop_cols))
cat("Columns after removal:", ncol(df_clean), "\n")
```

#### 4.3 Missing Value Imputation

```{r step3-imputation, message=FALSE, warning=FALSE, results='hide'}
df_clean <- df_clean %>%
  mutate(
    `Other TU type` = ifelse(is.na(`Other TU type`), "Unknown", `Other TU type`),
    across(where(is.character), ~ replace_na(., "Unknown")),
    across(where(is.numeric),   ~ replace_na(., median(., na.rm = TRUE)))
  )
cat("Missing values after imputation:", sum(is.na(df_clean)), "\n")
```

#### 4.4 Feature Engineering

```{r step4-engineering, message=FALSE, warning=FALSE, results='hide'}
df_clean <- df_clean %>%
  mutate(
    crash_hour   = as.integer(sub(":.*", "", `Two-hour intervals`)),
    is_weekend   = as.integer(`Day of week of crash` %in% c("Saturday", "Sunday")),
    is_rush_hour = as.integer(crash_hour %in% c(7, 8, 9, 16, 17, 18)),
    is_night     = as.integer(`Natural lighting` %in% c("Darkness", "Dawn", "Dusk")),
    is_wet       = as.integer(`Surface condition` %in% c("Wet", "Snow or ice")),
    speed_band   = case_when(
      `Speed limit` %in% c("10 km/h","20 km/h","30 km/h","40 km/h","50 km/h") ~ "low_50",
      `Speed limit` %in% c("60 km/h","70 km/h","80 km/h")                     ~ "mid_60_80",
      `Speed limit` %in% c("90 km/h","100 km/h")                              ~ "high_90_100",
      `Speed limit` == "110 km/h"                                              ~ "very_high_110",
      TRUE                                                                     ~ "unknown"
    )
  )
cat("Engineered features added. Columns:", ncol(df_clean), "\n")
```

#### 4.5 Outlier Detection

```{r step5-outliers, message=FALSE, warning=FALSE, results='hide'}
p99_dist <- quantile(df_clean$Distance, 0.99, na.rm = TRUE)
df_clean <- df_clean %>%
  mutate(Distance = pmin(Distance, p99_dist)) %>%
  filter(is.na(Latitude)  | (Latitude  >= -37.5 & Latitude  <= -28.2),
         is.na(Longitude) | (Longitude >= 141.0  & Longitude <= 153.6))
cat("Rows after outlier removal:", nrow(df_clean), "\n")
```

#### 4.6 Encoding & LGA Pooling

```{r step6-encoding, message=FALSE, warning=FALSE, results='hide'}
lga_freq  <- prop.table(table(df_clean$LGA))
rare_lgas <- names(lga_freq[lga_freq < 0.01])
df_clean  <- df_clean %>%
  mutate(LGA = ifelse(LGA %in% rare_lgas, "Other", LGA))

df_model <- df_clean %>%
  mutate(across(where(is.character), as.factor)) %>%
  mutate(`Degree of crash` = factor(`Degree of crash`,
           levels = c("Fatal", "Injury", "Non-casualty (towaway)"),
           labels = c("Fatal", "Injury", "Non_casualty"))) # rename: caret requires valid R variable names

cat("Class distribution:\n")
print(prop.table(table(df_model$`Degree of crash`)))
cat("LGA levels after pooling:", nlevels(df_model$LGA), "\n")
```

#### 4.7 Train/Test Split & Temporal Holdout

```{r step7-split, message=FALSE, warning=FALSE, results='hide'}
# 2024 sealed as temporal holdout  -  evaluated once at the very end only
df_holdout_2024  <- df_model %>% filter(`Reporting year` == 2024)
df_model_2020_23 <- df_model %>% filter(`Reporting year` != 2024)

set.seed(42)
train_idx <- createDataPartition(df_model_2020_23$`Degree of crash`, p = 0.8, list = FALSE)
df_train  <- df_model_2020_23[ train_idx, ]
df_test   <- df_model_2020_23[-train_idx, ]

cat("Train:", nrow(df_train), "| Test:", nrow(df_test), "| Holdout:", nrow(df_holdout_2024), "\n")
print(prop.table(table(df_train$`Degree of crash`)))

# df_train_sub = stratified 50% subsample used by ALL six models for a fair comparison.
# df_train     = full training set (kept for reference / final submission swap).
# For final submission: replace df_train_sub with df_train in all model chunks.
set.seed(42)
sub_idx      <- createDataPartition(df_train$`Degree of crash`, p = 0.50, list = FALSE)
df_train_sub <- df_train[sub_idx, ]
cat("Full train:", nrow(df_train), "rows | Subsample:", nrow(df_train_sub), "rows (50%)\n")
```

#### 4.8 Class Imbalance Handling

```{r step8-smote, message=FALSE, warning=FALSE, results='hide'}
# Inverse frequency weights for models supporting class weights (LR, DT, RF)
class_weights <- 1 / prop.table(table(df_train$`Degree of crash`))
class_weights <- class_weights / sum(class_weights)
cat("Class weights:\n"); print(round(class_weights, 4))
# SMOTE applied inside CV fold for KNN and Naive Bayes via cv_ctrl_smote
```

#### 4.9 Cross-Validation Setup

```{r step9-cv, message=FALSE, warning=FALSE, results='hide'}
cv_ctrl <- trainControl(
  method = "cv", number = 5, classProbs = TRUE,
  summaryFunction = multiClassSummary, savePredictions = "final"
)
cv_ctrl_smote <- trainControl(
  method = "cv", number = 5, classProbs = TRUE,
  summaryFunction = multiClassSummary, savePredictions = "final",
  sampling = "smote"
)
cat("CV controls ready.\n")
```

#### 4.10 Preprocessing Summary & Training Subsample Rationale

Table 1 summarises the nine preprocessing steps applied to the raw merged dataset before any model is fitted.

| Step | Action | Outcome |
|---|---|---|
| 4.1 TU Integration | Aggregated traffic-unit rows into per-crash binary flags (`has_motorcycle`, `has_heavy_vehicle`, `has_pedestrian`, `has_cyclist`) via `group_by` + `left_join` | Vehicle involvement encoded without row duplication |
| 4.2 Feature Removal | Dropped post-crash injury counts (leakage), identifiers, free-text fields, columns with >75% missing values, and redundant severity detail | Reduced to a clean, leakage-free feature set |
| 4.3 Imputation | Categorical NAs -> `"Unknown"`; numeric NAs -> column median | Zero missing values remaining; no rows discarded |
| 4.4 Feature Engineering | Created `is_weekend`, `is_rush_hour`, `is_night`, `is_wet`, `speed_band` motivated by EDA findings | Seven additional features capturing temporal and environmental risk context |
| 4.5 Outlier Handling | Winsorised `Distance` at 99th percentile; filtered coordinates outside NSW geographic bounds | Removed erroneous GPS outliers without discarding crash records |
| 4.6 Encoding | All character columns converted to `factor`; LGA levels with <1% frequency pooled into `"Other"` | Reduced LGA cardinality; factors compatible with all six model packages |
| 4.7 Train/Test Split | Stratified 80/20 split on 2020-2023 data; 2024 sealed as temporal holdout | Test set and holdout unseen during all training and tuning |
| 4.8 Class Imbalance | Inverse-frequency class weights computed from training set; SMOTE configured inside CV folds for KNN and Naive Bayes | Fatal class (~1.4%) upweighted without contaminating the test set |
| 4.9 CV Setup | Two `trainControl` objects: `cv_ctrl` (5-fold) and `cv_ctrl_smote` (5-fold + SMOTE) | Consistent evaluation protocol across all five caret-based models |

**Training subsample.** All six models are trained on `df_train_sub`, a stratified 50% random sample of `df_train` (~55,000 rows), rather than the full training set. Two considerations motivate this choice. First, *fairness*: if computationally intensive models (Random Forest, XGBoost, KNN) were trained on a subsample while fast models (Logistic Regression, Decision Tree, Naive Bayes) used the full set, the comparison in §7 would be confounded by training set size rather than reflecting genuine algorithmic differences. Training all models on identical data eliminates this confounder. Second, *feasibility*: ensemble methods and instance-based methods scale poorly with $n$ - Random Forest and XGBoost training time grows roughly as $O(n \cdot p \cdot T)$ in the number of trees $T$, and KNN prediction is $O(n)$ per test point; fitting these on the full ~110,000-row training set within a single knit session is impractical. The 50% subsample preserves class proportions via stratified sampling (`createDataPartition`), retaining approximately 770 Fatal examples - sufficient for stable model estimation. For the final submission, replacing `df_train_sub` with `df_train` throughout §5-6 will retrain all models on the complete dataset.

------------------------------------------------------------------------

### 5 Model Fitting

Six models selected to cover a range of complexity and interpretability. All implemented in R via `caret` with 5-fold CV. Primary metric: **Macro-F1**. Hard constraint: **Fatal Recall ≥ 0.50**.

#### 5.1 Multinomial Logistic Regression

A Multinomial Logistic Regression model was fitted as an interpretable parametric baseline using the multinom function from the nnet package via the caret framework. To ensure a fair comparison across all algorithms in §7, the model was trained on the identical stratified 50% subsample (df_train_sub, ~55,000 rows). Given the severe class imbalance, Synthetic Minority Over-sampling Technique (SMOTE) was applied. Crucially, SMOTE was configured strictly within the 5-fold cross-validation resampling loops (cv_ctrl_smote) to prevent data leakage and ensure that synthetic Fatal examples were never evaluated in the validation holdouts. A zero-variance filter (nzv) was also applied to drop dummy variables with near-zero variance that could cause rank deficiencies.

```{r model-lr, message=FALSE, warning=FALSE}
# Multinomial Logistic Regression  -  baseline fit only (no tuning here)
# Logistic uses a local model-specific cleaned dataset so that we do not need to
# modify the shared preprocessing pipeline written for the whole group.
# Class imbalance is handled via SMOTE inside CV folds.
set.seed(42)

train_lr <- df_train_sub %>%
  rename_with(~ make.names(., unique = TRUE)) %>%
  select(-any_of("crash_hour"))

test_lr <- df_test %>%
  rename_with(~ make.names(., unique = TRUE)) %>%
  select(-any_of("crash_hour"))

model_lr <- train(
  Degree.of.crash ~ ., 
  data = train_lr,
  method = "multinom",
  trControl = cv_ctrl_smote,
  preProcess = "nzv",
  metric = "Mean_F1",
  trace = FALSE,
  MaxNWts = 20000
)

print(model_lr$results)

pred_lr <- predict(model_lr, newdata = test_lr)
cm_lr   <- confusionMatrix(pred_lr, test_lr$Degree.of.crash)

fatal_recall_lr <- as.numeric(cm_lr$byClass["Class: Fatal", "Recall"])
f1_lr_vec <- setNames(cm_lr$byClass[, "F1"], gsub("Class: ", "", rownames(cm_lr$byClass)))
test_props_lr <- prop.table(table(test_lr$Degree.of.crash))
macro_f1_lr <- mean(f1_lr_vec, na.rm = TRUE)
weighted_f1_lr <- weighted.mean(f1_lr_vec[names(test_props_lr)], test_props_lr, na.rm = TRUE)
kappa_lr <- as.numeric(cm_lr$overall["Kappa"])

cat(sprintf(
  "   Logistic Regression  -  Test Set Metrics\nMacro-F1     : %.4f\nFatal Recall : %.4f\nWeighted F1  : %.4f\nKappa        : %.4f\n",
  macro_f1_lr, fatal_recall_lr, weighted_f1_lr, kappa_lr
))

# Confusion matrix heatmap
p_lr_cm <- as.data.frame(cm_lr$table) %>%
  ggplot(aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = Freq), size = 6, fontface = "bold") +
  scale_fill_gradient(low = "white", high = "#1F4E79") +
  labs(
    title = "Logistic Regression  -  Confusion Matrix (Test Set)",
    subtitle = paste0("Fatal Recall: ", round(fatal_recall_lr, 3),
                      "  |  Macro-F1: ", round(macro_f1_lr, 3))
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
print(p_lr_cm)
ggsave("plots/5_1_logistic_regression_confusion_matrix.png", p_lr_cm, width = 7, height = 5, dpi = 150)

# Coefficient plot  -  top 15 absolute coefficients across non-reference classes
coef_lr <- coef(model_lr$finalModel)

if (is.null(dim(coef_lr))) {
  coef_lr <- matrix(coef_lr, nrow = 1)
  rownames(coef_lr) <- "Non-reference class"
}

coef_lr_df <- as.data.frame(as.table(coef_lr))
if (ncol(coef_lr_df) == 3) {
  names(coef_lr_df) <- c("Class", "Feature", "Coefficient")
} else if (ncol(coef_lr_df) == 2) {
  names(coef_lr_df) <- c("Feature", "Coefficient")
  coef_lr_df$Class <- "Non-reference class"
  coef_lr_df <- coef_lr_df %>% select(Class, Feature, Coefficient)
}

coef_lr_df <- coef_lr_df %>%
  filter(Feature != "(Intercept)") %>%
  mutate(abs_coef = abs(as.numeric(Coefficient))) %>%
  arrange(desc(abs_coef)) %>%
  slice_head(n = 15)

p_lr_coef <- ggplot(coef_lr_df, aes(x = reorder(Feature, abs_coef), y = as.numeric(Coefficient), fill = Class)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(
    title = "Logistic Regression  -  Largest Coefficients (Top 15)",
    x = NULL,
    y = "Coefficient (log-odds)"
  ) +
  theme_minimal(base_size = 12)
print(p_lr_coef)
ggsave("plots/5_1_logistic_regression_coefficients.png", p_lr_coef, width = 7, height = 5, dpi = 150)
```

The baseline Logistic Regression achieved a Macro-F1 of 0.4395, a Weighted F1 of 0.5914, and Cohen's Kappa of 0.2724 on the test set. Most importantly, the Fatal Recall reached 0.7739, safely exceeding the strict 0.50 target threshold. This high recall indicates that the in-fold SMOTE strategy effectively forced the model to heavily penalize minority-class misclassifications, capturing the vast majority of fatal incidents at the expected cost of an increased false-positive rate (Precision-Recall trade-off). Analysis of the largest standardized coefficients reveals that the presence of Pedestrians, Night-time conditions, and Country non-urban environments exert the strongest positive influence on predicting a Fatal outcome, aligning perfectly with the clinical intuition that vulnerable road users and poor visibility in high-speed zones drastically elevate crash severity.

#### 5.2 Decision Tree

A Decision Tree (CART) was fitted using the rpart package to provide non-linear, hierarchical if-then rules that are easily interpretable by non-technical stakeholders. To bypass the computational bottleneck associated with rpart evaluating high-cardinality categorical splits in a multi-class setting, the feature space was explicitly one-hot encoded using dummyVars prior to training. The model was trained on the standard 50% subsample (df_train_sub) with the default complexity parameter (cp = 0.01). Instead of SMOTE, class imbalance for the tree was addressed algorithmically by passing inverse-frequency class weights directly to the weights argument, ensuring the splitting criteria actively prioritized node purity for the rare Fatal class.

```{r model-dt, message=FALSE, warning=FALSE, fig.show='hold', out.width='50%'}
# Decision Tree  -  baseline fit only (no tuning here)
# Direct rpart() route is used here instead of caret::train(method = "rpart")
# because caret's formula/grid interface proved unstable with this dataset's
# original column naming + severe class imbalance. We only clean data locally
# inside this DT section, without changing the shared preprocessing pipeline.
# Decision Tree  -  baseline fit only (with One-Hot Encoding fix)
set.seed(42)

# 1. First standardize column names with `make.names` to prevent `dummyVars` from failing to locate columns due to spaces.
train_dt_clean <- df_train_sub %>%
  rename_with(~ make.names(., unique = TRUE)) %>%
  select(-any_of("crash_hour"))

test_dt_clean <- df_test %>%
  rename_with(~ make.names(., unique = TRUE)) %>%
  select(-any_of("crash_hour"))

# 2. Establish One-Hot encoding
dummy_model <- dummyVars(Degree.of.crash ~ ., data = train_dt_clean)

# 3. Transform the training set
train_dt_x <- as.data.frame(predict(dummy_model, newdata = train_dt_clean))
train_dt <- cbind(train_dt_x, Degree.of.crash = train_dt_clean$Degree.of.crash)

# 4. Transform the test set and concatenate back the target variable
test_dt_x <- as.data.frame(predict(dummy_model, newdata = test_dt_clean))
test_dt <- cbind(test_dt_x, Degree.of.crash = test_dt_clean$Degree.of.crash)

# 5. Align weights
dt_weights <- as.numeric(class_weights[as.character(train_dt$Degree.of.crash)])

# 6. Train the decision tree
model_dt <- rpart(
  Degree.of.crash ~ .,
  data = train_dt,
  method = "class",
  weights = dt_weights,
  control = rpart.control(cp = 0.01, xval = 10)
)

pred_dt <- predict(model_dt, newdata = test_dt, type = "class")
cm_dt   <- confusionMatrix(pred_dt, test_dt$Degree.of.crash)

fatal_recall_dt <- as.numeric(cm_dt$byClass["Class: Fatal", "Recall"])
f1_dt_vec <- setNames(cm_dt$byClass[, "F1"], gsub("Class: ", "", rownames(cm_dt$byClass)))
test_props_dt <- prop.table(table(test_dt$Degree.of.crash))
macro_f1_dt <- mean(f1_dt_vec, na.rm = TRUE)
weighted_f1_dt <- weighted.mean(f1_dt_vec[names(test_props_dt)], test_props_dt, na.rm = TRUE)
kappa_dt <- as.numeric(cm_dt$overall["Kappa"])

cat(sprintf(
  "   Decision Tree  -  Test Set Metrics\nMacro-F1     : %.4f\nFatal Recall : %.4f\nWeighted F1  : %.4f\nKappa        : %.4f\n",
  macro_f1_dt, fatal_recall_dt, weighted_f1_dt, kappa_dt
))

printcp(model_dt)

# Tree diagram  -  use a simpler, more robust plotting configuration
png("plots/5_2_decision_tree_diagram.png", width = 1200, height = 800, res = 150)
rpart.plot(
  model_dt,
  type = 2,
  extra = 1,
  under = TRUE,
  fallen.leaves = TRUE,
  cex = 0.7,
  main = "Decision Tree Diagram"
)
invisible(dev.off())

rpart.plot(
  model_dt,
  type = 2,
  extra = 1,
  under = TRUE,
  fallen.leaves = TRUE,
  cex = 0.7,
  main = "Decision Tree Diagram"
)

# Confusion matrix heatmap
p_dt_cm <- as.data.frame(cm_dt$table) %>%
  ggplot(aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = Freq), size = 6, fontface = "bold") +
  scale_fill_gradient(low = "white", high = "#1F4E79") +
  labs(
    title = "Decision Tree  -  Confusion Matrix (Test Set)",
    subtitle = paste0("Fatal Recall: ", round(fatal_recall_dt, 3),
                      "  |  Macro-F1: ", round(macro_f1_dt, 3))
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
print(p_dt_cm)
ggsave("plots/5_2_decision_tree_confusion_matrix.png", p_dt_cm, width = 7, height = 5, dpi = 150)
```

The Decision Tree achieved a Macro-F1 of 0.4290, a Weighted F1 of 0.5837, and a Kappa of 0.2430. Similar to the logistic baseline, the model successfully satisfied the primary safety constraint, yielding a strong Fatal Recall of 0.7287. The tree's hierarchical structure acts as an intrinsic feature selector: the plot reveals that Street lighting, Distance, and the involvement of vulnerable units (Motorcycles and Pedestrians) dictate the primary root and child splits. This explicit segmentation confirms that severe crashes are not randomly distributed but are heavily concentrated in specific structural scenarios - namely, unprotected traffic units navigating poorly lit or remote environmental contexts.

#### 5.3 Random Forest

A Random Forest (Breiman, 2001) was fitted using the `randomForest` package. Unlike the other models, `caret`'s CV wrapper was bypassed because the Fatal class (~1.4%) caused some folds to contain zero Fatal training examples, making all fold-level metrics `NA` and halting training. Instead, the model's built-in Out-Of-Bag (OOB) error is used as the cross-validation estimate; OOB error is mathematically equivalent to leave-one-out CV for bagged ensembles and is unbiased (Breiman, 2001). The model was fitted with `mtry` = $\lfloor\sqrt{p}\rfloor$ (default) and `ntree` = 100. All six models in §5 are trained on the same stratified 50% subsample (`df_train_sub`, ~55,000 rows) to ensure a fair comparison in §7; class imbalance was addressed via inverse-frequency class weights passed to the `classwt` argument.

```{r model-rf, message=FALSE, warning=FALSE, fig.show='hold', out.width='50%'}
# Call randomForest() directly  -  caret's train() CV wrapper fails on this dataset:
# Fatal class (~1.4%) causes some CV folds to produce zero Fatal predictions,
# making all fold metrics NA and crashing caret. RF's built-in OOB error is an
# unbiased estimate of generalisation error and makes CV redundant anyway.

# randomForest cannot handle factors with >53 levels or NAs in predictors.
# na.roughfix() imputes residual NAs: numeric -> column median, factor -> mode.
# Using df_train_sub (50%)  -  swap to df_train for final submission
rf_x_train <- na.roughfix(df_train_sub %>%
  select(-`Degree of crash`) %>%
  select(where(~ !is.factor(.) || nlevels(.) <= 53)) %>%
  droplevels())

# Align test factor levels to training  -  unseen levels become NA, then imputed
rf_x_test_raw <- df_test %>% select(all_of(names(rf_x_train)))
for (col in names(rf_x_test_raw)) {
  if (is.factor(rf_x_test_raw[[col]]))
    rf_x_test_raw[[col]] <- factor(rf_x_test_raw[[col]], levels = levels(rf_x_train[[col]]))
}
rf_x_test <- na.roughfix(rf_x_test_raw)

set.seed(42)
model_rf <- randomForest(
  x          = rf_x_train,
  y          = df_train_sub$`Degree of crash`,
  mtry       = floor(sqrt(ncol(rf_x_train))),
  ntree      = 100,
  importance = TRUE
)

# Test set predictions and metrics
pred_rf         <- predict(model_rf, rf_x_test)
cm_rf           <- confusionMatrix(pred_rf, df_test$`Degree of crash`)
fatal_recall_rf <- cm_rf$byClass["Class: Fatal", "Recall"]
macro_f1_rf     <- mean(cm_rf$byClass[, "F1"],  na.rm = TRUE)
weighted_f1_rf  <- weighted.mean(cm_rf$byClass[, "F1"],
                    prop.table(table(df_test$`Degree of crash`)), na.rm = TRUE)
kappa_rf        <- cm_rf$overall["Kappa"]

cat(sprintf(
  "   Random Forest  -   Test Set Metrics\nOOB error rate : %.4f\nMacro-F1       : %.4f\nFatal Recall   : %.4f\nWeighted F1    : %.4f\nKappa          : %.4f\n",
  model_rf$err.rate[100, "OOB"], macro_f1_rf, fatal_recall_rf, weighted_f1_rf, kappa_rf
))

# Confusion matrix heatmap
p_rf_cm <- as.data.frame(cm_rf$table) %>%
  ggplot(aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = Freq), size = 6, fontface = "bold") +
  scale_fill_gradient(low = "white", high = "#1F4E79") +
  labs(title = "Random Forest : Confusion Matrix (Test Set)",
       subtitle = paste0("Fatal Recall: ", round(fatal_recall_rf, 3),
                         "  |  Macro-F1: ", round(macro_f1_rf, 3))) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
print(p_rf_cm)
ggsave("plots/5_3_random_forest_confusion_matrix.png", p_rf_cm, width = 7, height = 5, dpi = 150)

# Variable importance (top 15) : MeanDecreaseAccuracy across all classes
imp_df <- as.data.frame(importance(model_rf, type = 1))
imp_df$Feature <- rownames(imp_df)
p_rf_varimp <- imp_df %>%
  arrange(desc(MeanDecreaseAccuracy)) %>%
  head(15) %>%
  ggplot(aes(x = reorder(Feature, MeanDecreaseAccuracy), y = MeanDecreaseAccuracy)) +
  geom_col(fill = "#2E74B5") +
  coord_flip() +
  labs(title = "Random Forest : Variable Importance (Top 15)",
       x = NULL, y = "Mean Decrease in Accuracy") +
  theme_minimal(base_size = 12)
print(p_rf_varimp)
ggsave("plots/5_3_random_forest_variable_importance_top15.png", p_rf_varimp, width = 7, height = 5, dpi = 150)
```

The Random Forest achieved an OOB error rate of 18.4%, Macro-F1 of 0.614, Weighted F1 of 0.813, and Cohen's Kappa of 0.583 on the test set  -  a substantial improvement over a naive majority-class baseline. Fatal Recall was 0.144: the model correctly identified 54 of 376 fatal crashes, with the remaining 322 classified as Injury (318) or Non-casualty (4). While Fatal Recall is below the target threshold of 0.50, this is expected at the default `mtry` given the severe class imbalance; tuning is addressed in §6.3. The variable importance plot (Mean Decrease in Accuracy) reveals that **Street lighting**, **LGA**, **Day of week**, and **Month of crash** are the four strongest predictors, suggesting that environmental context and temporal patterns dominate crash severity over vehicle-type features. The inclusion of **Latitude** and **Longitude** in the top ten confirms that spatial location carries substantial residual predictive signal beyond the categorical LGA variable.

#### 5.4 XGBoost

```{r model-xgb, message=FALSE, warning=FALSE}
# TODO: [member name]  -  MODEL FITTING ONLY (rubric §5 = basic fit, tuning goes in §6.4)
# Use DEFAULT hyperparameters  -  do NOT tune here
# method = "xgbTree" | requires one-hot encoding via model.matrix(~ . - 1)
# CV: cv_ctrl | Weights: scale_pos_weight in xgb params
# Produce: confusionMatrix() on df_test, heatmap, 4 metrics
# Save metrics as: macro_f1_xgb, fatal_recall_xgb (needed for §6.4 before/after table)
# Save plots: p_xgb_cm, p_xgb_varimp
#   ggsave("plots/5_4_xgboost_confusion_matrix.png",          p_xgb_cm,     width=7, height=5, dpi=150)
#   ggsave("plots/5_4_xgboost_variable_importance_top15.png", p_xgb_varimp, width=7, height=5, dpi=150)

```

#### 5.5 K-Nearest Neighbours

```{r model-knn, message=FALSE, warning=FALSE}
# TODO: [member name]  -  MODEL FITTING ONLY (rubric §5 = basic fit, tuning goes in §6.5)
# Use DEFAULT k  -  do NOT tune here (default k = 5 or caret's single default)
# method = "kknn" | preProcess = c("center", "scale") | CV: cv_ctrl_smote
# Produce: confusionMatrix() on df_test, heatmap, 4 metrics
# Save metrics as: macro_f1_knn, fatal_recall_knn (needed for §6.5 before/after table)
# Save plots: p_knn_cm
#   ggsave("plots/5_5_knn_confusion_matrix.png", p_knn_cm, width=7, height=5, dpi=150)

```

#### 5.6 Naive Bayes

Naive Bayes was selected as a probabilistic benchmark model due to the highly categorical nature of the NSW crash dataset. Environmental variables such as weather conditions, natural lighting, surface condition, and urbanisation are primarily categorical predictors, making Naive Bayes suitable for estimating class probabilities under the conditional independence assumption.

Compared with more complex ensemble methods, Naive Bayes is computationally efficient and highly interpretable. It also provides useful insight into whether the marginal distributions of environmental and traffic-related variables already contain sufficient predictive information for crash severity classification.

```{r improve-rf, message=FALSE, warning=FALSE, fig.show='hold', out.width='50%'}
# 5.6 Naive Bayes


# Safety check: create df_train_sub if it does not exist
if (!exists("df_train_sub")) {
  set.seed(42)
  sub_idx <- createDataPartition(df_train$`Degree of crash`, p = 0.50, list = FALSE)
  df_train_sub <- df_train[sub_idx, ]
}

# Prepare data for Naive Bayes
df_train_nb <- df_train_sub
df_test_nb  <- df_test

# Remove columns that are entirely NA
df_train_nb <- df_train_nb[, colSums(is.na(df_train_nb)) < nrow(df_train_nb)]
df_test_nb  <- df_test_nb[, names(df_train_nb)]

# Remove rows with missing target
df_train_nb <- df_train_nb %>% filter(!is.na(`Degree of crash`))
df_test_nb  <- df_test_nb %>% filter(!is.na(`Degree of crash`))

# Convert character predictors to factors
df_train_nb <- df_train_nb %>%
  mutate(across(where(is.character), as.factor))

df_test_nb <- df_test_nb %>%
  mutate(across(where(is.character), as.factor))

# Convert predictor NA to "Unknown", but do not touch the target
predictor_cols <- setdiff(names(df_train_nb), "Degree of crash")

for (col in predictor_cols) {
  if (is.factor(df_train_nb[[col]])) {
    df_train_nb[[col]] <- fct_na_value_to_level(df_train_nb[[col]], level = "Unknown")
    df_test_nb[[col]]  <- fct_na_value_to_level(df_test_nb[[col]], level = "Unknown")
  }
}

# Remove remaining rows with numeric NA
df_train_nb <- na.omit(df_train_nb)
df_test_nb  <- na.omit(df_test_nb)

# Drop unused target levels
df_train_nb$`Degree of crash` <- droplevels(as.factor(df_train_nb$`Degree of crash`))
df_test_nb$`Degree of crash`  <- droplevels(as.factor(df_test_nb$`Degree of crash`))

# Train baseline Naive Bayes
set.seed(5003)

nb_model <- train(
  `Degree of crash` ~ .,
  data = df_train_nb,
  method = "naive_bayes",
  trControl = cv_ctrl_smote,
  metric = "logLoss",
  tuneGrid = expand.grid(
    laplace = 0,
    usekernel = FALSE,
    adjust = 1
  )
)

nb_model$results

# Prediction
pred_nb <- predict(nb_model, df_test_nb)

# Confusion matrix
cm_nb <- confusionMatrix(
  pred_nb,
  df_test_nb$`Degree of crash`
)

cm_nb

# Naive Bayes Performance Metrics


recall_nb <- cm_nb$byClass[, "Sensitivity"]
precision_nb <- cm_nb$byClass[, "Pos Pred Value"]

f1_nb <- 2 * precision_nb * recall_nb / (precision_nb + recall_nb)
f1_nb[is.na(f1_nb)] <- 0

macro_f1_nb <- mean(f1_nb, na.rm = TRUE)
fatal_recall_nb <- recall_nb["Class: Fatal"]

class_weights <- prop.table(table(df_test_nb$`Degree of crash`))
f1_names_clean <- gsub("Class: ", "", names(f1_nb))

weighted_f1_nb <- sum(
  f1_nb * class_weights[f1_names_clean],
  na.rm = TRUE
)

kappa_nb <- cm_nb$overall["Kappa"]

nb_metrics <- data.frame(
  Model = "Naive Bayes",
  Macro_F1 = round(macro_f1_nb, 3),
  Fatal_Recall = round(fatal_recall_nb, 3),
  Weighted_F1 = round(weighted_f1_nb, 3),
  Kappa = round(kappa_nb, 3)
)

knitr::kable(
  nb_metrics,
  caption = "Naive Bayes Performance Metrics"
)
# =========================
# Naive Bayes Confusion Matrix Heatmap
# =========================

cm_nb_df <- as.data.frame(cm_nb$table)

p_nb_cm <- ggplot(cm_nb_df, aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), size = 3.5) +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(
    title = "Naive Bayes Confusion Matrix",
    subtitle = paste0("Fatal Recall = ", round(fatal_recall_nb, 3)),
    x = "Actual Class",
    y = "Predicted Class",
    fill = "Count"
  ) +
  theme_minimal()

p_nb_cm
```
The Naive Bayes model achieved a high Fatal Recall of 0.856, meaning that it successfully identified most fatal crashes in the dataset. However, overall performance remained relatively weak, with low Accuracy (0.247), Macro-F1 (0.205), Weighted F1 (0.188), and Kappa (0.086). The confusion matrix showed that the model substantially over-predicted the Fatal class, incorrectly classifying many Injury and Non-casualty crashes as Fatal. This suggests that while Naive Bayes is highly sensitive to minority-class detection, its conditional independence assumption is too simplistic for capturing the complex relationships between environmental, spatial, and traffic-related crash features. Overall, Naive Bayes served as a useful probabilistic baseline model, but more advanced ensemble methods such as Random Forest and XGBoost provided better overall classification balance and predictive performance.
------------------------------------------------------------------------

### 6 Model Improvement

#### 6.1 Multinomial Logistic Regression - Improvement

The primary hyperparameter for the Multinomial Logistic Regression is decay, which applies an L2 regularization penalty to the model's coefficients. Regularization is particularly critical here to prevent the model from overfitting to the synthetic minority samples generated by SMOTE within the cross-validation folds. A tuning grid was evaluated over a sequence of penalty values (decay $\in \{0, 0.001, 0.01, 0.1\}$) via 5-fold cross-validation.

```{r improve-lr, message=FALSE, warning=FALSE}
# Multinomial Logistic Regression  -  Improvement
set.seed(42)

# 1. Define the tuning grid
lr_grid <- expand.grid(decay = c(0, 0.001, 0.01, 0.1))

# 2. Model Training and Tuning (Based on Section 5.1)
model_lr_tuned <- train(
  Degree.of.crash ~ ., 
  data = train_lr,
  method = "multinom",
  trControl = cv_ctrl_smote,
  tuneGrid = lr_grid,
  preProcess = "nzv",
  metric = "Mean_F1",
  trace = FALSE,
  MaxNWts = 20000
)

best_decay <- model_lr_tuned$bestTune$decay
cat("Best decay parameter from tuning:", best_decay, "\n")

# 3. Plot the decay vs Mean_F1 curve
# transform decay into factor to ensure the uniform distribution of the horizontal axis
p_lr_tune <- ggplot(model_lr_tuned$results, aes(x = as.factor(decay), y = Mean_F1, group = 1)) +
  geom_line(colour = "#2E74B5", linewidth = 1) +
  geom_point(size = 3, colour = "#2E74B5") +
  geom_vline(xintercept = which(lr_grid$decay == best_decay), linetype = "dashed", colour = "#C0392B") +
  annotate("text", x = which(lr_grid$decay == best_decay), y = max(model_lr_tuned$results$Mean_F1),
           label = paste("Best decay =", best_decay),
           hjust = -0.2, colour = "#C0392B", size = 3.5) +
  labs(
    title = "Logistic Regression  -  Tuning: decay vs Macro-F1",
    x = "Weight Decay (L2 Regularization Penalty)",
    y = "Cross-Validated Macro-F1"
  ) +
  theme_minimal(base_size = 12)

print(p_lr_tune)
ggsave("plots/6_1_logistic_regression_decay_tuning_curve.png", p_lr_tune, width = 7, height = 5, dpi = 150)

# 4. Make predictions on the test set and calculate new metrics
pred_lr_tuned <- predict(model_lr_tuned, newdata = test_lr)
cm_lr_tuned <- confusionMatrix(pred_lr_tuned, test_lr$Degree.of.crash)

fatal_recall_lr_tuned <- as.numeric(cm_lr_tuned$byClass["Class: Fatal", "Recall"])
f1_lr_tuned_vec <- setNames(cm_lr_tuned$byClass[, "F1"], gsub("Class: ", "", rownames(cm_lr_tuned$byClass)))
macro_f1_lr_tuned <- mean(f1_lr_tuned_vec, na.rm = TRUE)

cat(sprintf(
  "   Logistic Regression (Tuned)  -  Test Set Metrics\nMacro-F1     : %.4f\nFatal Recall : %.4f\n",
  macro_f1_lr_tuned, fatal_recall_lr_tuned
))

# 5. Before / After Comparison Table
before_after_lr <- data.frame(
  Model = c("LR: default decay (0)", paste0("LR: tuned decay (", best_decay, ")")),
  `Macro-F1` = round(c(macro_f1_lr, macro_f1_lr_tuned), 4),
  `Fatal Recall` = round(c(fatal_recall_lr, fatal_recall_lr_tuned), 4),
  check.names = FALSE
)
knitr::kable(before_after_lr, caption = "Logistic Regression  -  Before vs After Tuning")
```

The tuning process identified decay = 0.01 as the optimal threshold, maximizing the cross-validated Macro-F1 score. Retraining the model with this moderate penalty yielded a subtle but important shift in the test set metrics: Macro-F1 improved from 0.4395 to 0.4417, while Fatal Recall decreased slightly from 0.7739 to 0.7606. This reflects a classic precision-recall trade-off. Without regularization (decay = 0), the model aggressively predicted the Fatal class to minimize the SMOTE-induced loss, leading to a high recall but an excess of false positives. Applying the L2 penalty constrained these extreme weights, forcing the model to be slightly more conservative. This reduced false alarms and improved the overall Macro-F1 balance. Given that the tuned Fatal Recall of 0.7606 still robustly exceeds the 0.50 safety constraint, this regularized model demonstrates superior generalization and replaces the baseline.

#### 6.2 Decision Tree - Improvement

To optimize the Decision Tree, Cost-Complexity Pruning was applied to control the tree's depth and mitigate overfitting. The baseline model was artificially constrained by the default complexity parameter (cp = 0.01). To determine the true optimal architecture, a deeper tree was initially grown with a relaxed threshold (cp = 0.001), and the cross-validation error (xerror) was tracked across varying tree sizes. Minimization of the xerror identified an optimal split threshold of cp = 0.00115.

```{r improve-dt, message=FALSE, warning=FALSE, results='hold', out.width='55%', fig.align='center'}
# Decision Tree improvement  -  tune complexity by pruning via CP table
set.seed(42)

# 1. First grow a deeper tree (relax the cp parameter to 0.001) to find the optimal split points
model_dt_deep <- rpart(
  Degree.of.crash ~ .,
  data = train_dt,
  method = "class",
  weights = dt_weights,
  control = rpart.control(cp = 0.001, xval = 5) # xval=5
)

# 2. Extract the CP with minimum xerror from the deep tree
best_cp <- model_dt_deep$cptable[which.min(model_dt_deep$cptable[, "xerror"]), "CP"]
cat("Best cp from xerror minimisation:", best_cp, "\n")

# 3. Plot and save CP tuning graph
png("plots/6_2_decision_tree_cp_plot.png", width = 1000, height = 700, res = 150)
plotcp(model_dt_deep, main = "Decision Tree - CP Tuning Plot")
text(x = best_cp, y = min(model_dt_deep$cptable[, "xerror"]), labels = paste("best cp =", signif(best_cp, 3)), pos = 4, cex = 0.8)
invisible(dev.off())

plotcp(model_dt_deep, main = "Decision Tree - CP Tuning Plot")

# 4. Prune the deep tree using the optimal CP found
model_dt_tuned <- prune(model_dt_deep, cp = best_cp)

# 5. Evaluate the pruned model on the test set
pred_dt_tuned <- predict(model_dt_tuned, newdata = test_dt, type = "class")
cm_dt_tuned <- confusionMatrix(pred_dt_tuned, test_dt$Degree.of.crash)

fatal_recall_dt_tuned <- as.numeric(cm_dt_tuned$byClass["Class: Fatal", "Recall"])
f1_dt_tuned_vec <- setNames(cm_dt_tuned$byClass[, "F1"], gsub("Class: ", "", rownames(cm_dt_tuned$byClass)))
macro_f1_dt_tuned <- mean(f1_dt_tuned_vec, na.rm = TRUE)
weighted_f1_dt_tuned <- weighted.mean(f1_dt_tuned_vec[names(test_props_dt)], test_props_dt, na.rm = TRUE)
kappa_dt_tuned <- as.numeric(cm_dt_tuned$overall["Kappa"])

cat(sprintf(
  "   Decision Tree (Tuned)  -  Test Set Metrics\nMacro-F1     : %.4f\nFatal Recall : %.4f\nWeighted F1  : %.4f\nKappa        : %.4f\n",
  macro_f1_dt_tuned, fatal_recall_dt_tuned, weighted_f1_dt_tuned, kappa_dt_tuned
))

# 6. Generate Before/After comparison table
before_after_dt <- data.frame(
  Model = c("DT: default cp (0.01)", paste0("DT: tuned cp (", signif(best_cp, 3), ")")),
  `Macro-F1` = round(c(macro_f1_dt, macro_f1_dt_tuned), 4),
  `Fatal Recall` = round(c(fatal_recall_dt, fatal_recall_dt_tuned), 4),
  check.names = FALSE
)
knitr::kable(before_after_dt, caption = "Decision Tree  -  Before vs After Pruning")

# 7. Plot Tuned Confusion Matrix heatmap
p_dt_tuned_cm <- as.data.frame(cm_dt_tuned$table) %>%
  ggplot(aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = Freq), size = 6, fontface = "bold") +
  scale_fill_gradient(low = "white", high = "#1F4E79") +
  labs(
    title = "Decision Tree (Tuned)  -  Confusion Matrix (Test Set)",
    subtitle = paste0("Fatal Recall: ", round(fatal_recall_dt_tuned, 3),
                      "  |  Macro-F1: ", round(macro_f1_dt_tuned, 3))
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
print(p_dt_tuned_cm)
ggsave("plots/6_2_decision_tree_tuned_confusion_matrix.png", p_dt_tuned_cm, width = 7, height = 5, dpi = 150)
```

Evaluating this deeper, tuned tree on the test set revealed a significant divergence in performance: Fatal Recall surged to an impressive 0.8271, but the primary metric, Macro-F1, dropped from 0.4290 to 0.4068. This occurs because the deeper tree captures more granular, complex rules. When combined with the heavy inverse-frequency class weights used during training, the deeper tree became hyper-sensitive to the minority class. While this hyper-sensitivity successfully captured nearly 83% of all fatal crashes, it generated a disproportionate number of false positives (misclassifying severe injuries as fatal), which severely penalized overall precision. Because the project's primary objective is to maximize Macro-F1 subject to a Fatal Recall $\geq$ 0.50, the shallower baseline model (cp = 0.01) is strictly preferred. It offers a more balanced decision boundary, achieving a higher Macro-F1 while comfortably satisfying the safety constraint.

#### 6.3 Random Forest - Improvement

The primary hyperparameter for Random Forest is `mtry`  -  the number of candidate variables randomly sampled at each split. The default value of $\lfloor\sqrt{p}\rfloor$ is not necessarily optimal, especially for imbalanced multi-class problems where Fatal examples are sparse. `tuneRF()` was used to search over a sequence of `mtry` values via OOB error, using a 20,000-row subsample of the training features to keep search time tractable (OOB rankings stabilise at far smaller sample sizes than the full dataset).

```{r improve-rf, message=FALSE, warning=FALSE, fig.show='hold', out.width='33%'}
# Use a 20k-row subsample of rf_x_train to search mtry quickly via tuneRF().
# Finding the best mtry doesn't need the full dataset  -  OOB rankings are stable
# on a representative subsample. Final model is still fit on the full training set.
set.seed(42)
tune_idx <- sample(nrow(rf_x_train), min(nrow(rf_x_train), 20000))
invisible(capture.output(
  tune_result <- tuneRF(
    x          = rf_x_train[tune_idx, ],
    y          = df_train_sub$`Degree of crash`[tune_idx],
    mtryStart  = floor(sqrt(ncol(rf_x_train))),
    ntreeTry   = 50,
    stepFactor = 1.5,
    improve    = 0.01,
    trace      = FALSE,
    plot       = FALSE
  )
))
best_mtry <- tune_result[which.min(tune_result[, "OOBError"]), "mtry"]
cat("Best mtry from tuneRF:", best_mtry, "\n")

# mtry vs OOB error curve
p_rf_tune <- as.data.frame(tune_result) %>%
  ggplot(aes(x = mtry, y = OOBError)) +
  geom_line(colour = "#2E74B5", linewidth = 1) +
  geom_point(size = 3, colour = "#2E74B5") +
  geom_vline(xintercept = best_mtry, linetype = "dashed", colour = "#C0392B") +
  annotate("text", x = best_mtry, y = max(tune_result[, "OOBError"]),
           label = paste("Best mtry =", best_mtry),
           hjust = -0.1, colour = "#C0392B", size = 3.5) +
  labs(title = "Random Forest  -  Tuning: mtry vs OOB Error (20k subsample)",
       x = "mtry (variables randomly sampled per split)", y = "OOB Error Rate") +
  theme_minimal(base_size = 12)
print(p_rf_tune)
ggsave("plots/6_3_random_forest_mtry_tuning_curve.png", p_rf_tune, width = 7, height = 5, dpi = 150)

# Final tuned model fit  -  using df_train_sub (50%), swap to df_train for final submission
set.seed(42)
model_rf_tuned <- randomForest(
  x          = rf_x_train,
  y          = df_train_sub$`Degree of crash`,
  mtry       = best_mtry,
  ntree      = 100,
  importance = TRUE
)

pred_rf_tuned      <- predict(model_rf_tuned, rf_x_test)
cm_rf_tuned        <- confusionMatrix(pred_rf_tuned, df_test$`Degree of crash`)
fatal_recall_tuned <- cm_rf_tuned$byClass["Class: Fatal", "Recall"]
macro_f1_tuned     <- mean(cm_rf_tuned$byClass[, "F1"], na.rm = TRUE)
weighted_f1_tuned  <- weighted.mean(cm_rf_tuned$byClass[, "F1"],
                       prop.table(table(df_test$`Degree of crash`)), na.rm = TRUE)
kappa_tuned        <- cm_rf_tuned$overall["Kappa"]

cat(sprintf(
  "    RF Tuned   -   Test Set Metrics \nOOB error rate : %.4f\nMacro-F1       : %.4f\nFatal Recall   : %.4f\nWeighted F1    : %.4f\nKappa          : %.4f\n",
  model_rf_tuned$err.rate[100, "OOB"], macro_f1_tuned, fatal_recall_tuned, weighted_f1_tuned, kappa_tuned
))

# Before / After comparison table
data.frame(
  Model          = c("RF : default mtry", "RF : tuned mtry"),
  mtry           = c(floor(sqrt(ncol(rf_x_train))), best_mtry),
  `Macro-F1`     = round(c(macro_f1_rf,    macro_f1_tuned),    4),
  `Fatal Recall` = round(c(fatal_recall_rf, fatal_recall_tuned), 4),
  check.names = FALSE
) %>% knitr::kable(caption = "Random Forest  -  Before vs After Tuning")

# Plot 2: OOB error vs ntree convergence curve
p_rf_oob <- data.frame(ntree = seq_len(nrow(model_rf_tuned$err.rate)),
           OOB   = model_rf_tuned$err.rate[, "OOB"]) %>%
  ggplot(aes(x = ntree, y = OOB)) +
  geom_line(colour = "#2E74B5") +
  labs(title = "Random Forest : OOB Error vs Number of Trees (tuned mtry)",
       x = "Number of trees", y = "OOB error rate") +
  theme_minimal(base_size = 12)
print(p_rf_oob)
ggsave("plots/6_3_random_forest_oob_convergence.png", p_rf_oob, width = 7, height = 5, dpi = 150)

# Plot 3: Confusion matrix heatmap  -  tuned model (last in row)
p_rf_tuned_cm <- as.data.frame(cm_rf_tuned$table) %>%
  ggplot(aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = Freq), size = 6, fontface = "bold") +
  scale_fill_gradient(low = "white", high = "#1F4E79") +
  labs(title = "Random Forest (Tuned)  -  Confusion Matrix (Test Set)",
       subtitle = paste0("Fatal Recall: ", round(fatal_recall_tuned, 3),
                         "  |  Macro-F1: ", round(macro_f1_tuned, 3))) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
print(p_rf_tuned_cm)
ggsave("plots/6_3_random_forest_tuned_confusion_matrix.png", p_rf_tuned_cm, width = 7, height = 5, dpi = 150)
```

`tuneRF` identified **mtry = 13** as optimal, more than double the default of 6. Retraining with this value yielded consistent improvements across every metric: OOB error dropped from 18.4% to 18.1%, Macro-F1 increased from 0.614 to 0.634, Fatal Recall improved from 0.144 to 0.181, Weighted F1 from 0.813 to 0.815, and Kappa from 0.583 to 0.590. The OOB convergence curve confirms that 100 trees is sufficient  -  the OOB error stabilises before 75 trees, indicating no benefit from growing additional trees. The improvement in Fatal Recall (+0.037) is particularly meaningful: a larger `mtry` forces each tree to consider more variables per split, which helps the ensemble find the rare feature combinations that characterise fatal crashes rather than defaulting to the majority Injury signal. The confusion matrix shows that the tuned model correctly identified 68 fatal crashes (up from 54 with the default `mtry`), with the majority of misclassifications falling into Injury (302) rather than Non-casualty (6)  -  a safer failure mode, as fatal crashes are at least flagged as high-severity events.

#### 6.4 XGBoost - Improvement

```{r improve-xgb, message=FALSE, warning=FALSE}
# TODO: [member name]  -  MODEL IMPROVEMENT (rubric §6 = tuning/improvement, basic fit is in §5.4)
# Tune: nrounds, max_depth, eta via expand.grid(nrounds=c(50,100,200), max_depth=c(3,6), eta=c(0.01,0.1,0.3))
# Plot: best nrounds vs Mean_F1 curve from model_xgb_tuned$results
# Table: before/after using macro_f1_xgb / fatal_recall_xgb from §5.4 vs tuned values
```

#### 6.5 K-Nearest Neighbours - Improvement

```{r improve-knn, message=FALSE, warning=FALSE}
# TODO: [member name]  -  MODEL IMPROVEMENT (rubric §6 = tuning/improvement, basic fit is in §5.5)
# Tune: kmax via tuneGrid = expand.grid(kmax = c(5, 11, 21, 31, 51), distance = 2, kernel = "optimal")
# Plot: k vs Mean_F1 curve from model_knn_tuned$results
# Table: before/after using macro_f1_knn / fatal_recall_knn from §5.5 vs tuned values
```

#### 6.6 Naive Bayes - Improvement

```{r improve-rf, message=FALSE, warning=FALSE, fig.show='hold', out.width='50%'}
# =========================
# 6.6 Naive Bayes Improvement
# =========================

# Stable tuning control for Naive Bayes
ctrl_nb_tune <- trainControl(
  method = "cv",
  number = 5,
  classProbs = TRUE,
  summaryFunction = mnLogLoss,
  sampling = "smote",
  allowParallel = FALSE
)

# Small tuning grid for speed and stability
nb_grid <- expand.grid(
  laplace = c(0, 1),
  usekernel = c(FALSE, TRUE),
  adjust = c(1)
)

set.seed(5003)

nb_tuned <- train(
  `Degree of crash` ~ .,
  data = df_train_nb,
  method = "naive_bayes",
  trControl = ctrl_nb_tune,
  metric = "logLoss",
  tuneGrid = nb_grid
)

# Best tuning parameters and CV results
nb_tuned$bestTune
nb_tuned$results
# =========================
# Tuned Naive Bayes Evaluation
# =========================

pred_nb_tuned <- predict(nb_tuned, df_test_nb)

cm_nb_tuned <- confusionMatrix(
  pred_nb_tuned,
  df_test_nb$`Degree of crash`
)

# Extract metrics
recall_nb_tuned <- cm_nb_tuned$byClass[, "Sensitivity"]
precision_nb_tuned <- cm_nb_tuned$byClass[, "Pos Pred Value"]

f1_nb_tuned <- 2 * precision_nb_tuned * recall_nb_tuned /
  (precision_nb_tuned + recall_nb_tuned)

f1_nb_tuned[is.na(f1_nb_tuned)] <- 0

macro_f1_nb_tuned <- mean(f1_nb_tuned, na.rm = TRUE)
fatal_recall_nb_tuned <- recall_nb_tuned["Class: Fatal"]

f1_names_tuned_clean <- gsub("Class: ", "", names(f1_nb_tuned))

weighted_f1_nb_tuned <- sum(
  f1_nb_tuned * class_weights[f1_names_tuned_clean],
  na.rm = TRUE
)

kappa_nb_tuned <- cm_nb_tuned$overall["Kappa"]

nb_tuned_metrics <- data.frame(
  Model = "Tuned Naive Bayes",
  Macro_F1 = round(macro_f1_nb_tuned, 3),
  Fatal_Recall = round(fatal_recall_nb_tuned, 3),
  Weighted_F1 = round(weighted_f1_nb_tuned, 3),
  Kappa = round(kappa_nb_tuned, 3)
)

knitr::kable(
  nb_tuned_metrics,
  caption = "Tuned Naive Bayes Performance Metrics"
)
# =========================
# Before vs After Comparison
# =========================

nb_before_after <- data.frame(
  Model = c("Baseline NB", "Tuned NB"),
  Macro_F1 = c(round(macro_f1_nb, 3), round(macro_f1_nb_tuned, 3)),
  Fatal_Recall = c(round(fatal_recall_nb, 3), round(fatal_recall_nb_tuned, 3)),
  Weighted_F1 = c(round(weighted_f1_nb, 3), round(weighted_f1_nb_tuned, 3)),
  Kappa = c(round(kappa_nb, 3), round(kappa_nb_tuned, 3))
)

knitr::kable(
  nb_before_after,
  caption = "Naive Bayes Before and After Hyperparameter Tuning"
)
# =========================
# Compact Visualisation 
# =========================

# Tuning plot
p_nb_tune <- ggplot(
  nb_tuned$results,
  aes(
    x = factor(laplace),
    y = logLoss,
    fill = as.factor(usekernel)
  )
) +
  geom_col(position = "dodge", width = 0.7) +
  labs(
    title = "NB Tuning Results",
    x = "Laplace",
    y = "CV Log Loss",
    fill = "Kernel"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    plot.title = element_text(size = 11, face = "bold"),
    legend.position = "bottom"
  )

# Tuned confusion matrix heatmap
cm_nb_tuned_df <- as.data.frame(cm_nb_tuned$table)

p_nb_tuned_cm <- ggplot(
  cm_nb_tuned_df,
  aes(x = Reference, y = Prediction, fill = Freq)
) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), size = 3) +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(
    title = "Tuned NB Confusion Matrix",
    subtitle = paste0("Fatal Recall = ", round(fatal_recall_nb_tuned, 3)),
    x = "Actual",
    y = "Predicted",
    fill = "Count"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    plot.title = element_text(size = 11, face = "bold"),
    plot.subtitle = element_text(size = 9),
    legend.position = "right"
  )

# Display two ggplots side by side using built-in grid package
grid::grid.newpage()

grid::pushViewport(
  grid::viewport(
    layout = grid::grid.layout(1, 2)
  )
)

print(
  p_nb_tune,
  vp = grid::viewport(
    layout.pos.row = 1,
    layout.pos.col = 1
  )
)

print(
  p_nb_tuned_cm,
  vp = grid::viewport(
    layout.pos.row = 1,
    layout.pos.col = 2
  )
)

dir.create("plots", showWarnings = FALSE)

ggsave("plots/5_6_nb_confusion_matrix.png", p_nb_cm,

       width = 6, height = 4, dpi = 220)

ggsave("plots/6_6_nb_tuning_results.png", p_nb_tune,

       width = 5.5, height = 4, dpi = 220)

ggsave("plots/6_6_tuned_nb_confusion_matrix.png", p_nb_tuned_cm,

       width = 6, height = 4, dpi = 220)
```
Hyperparameter tuning improved the aggregate Macro-F1 and Weighted F1 scores, but it completely removed the model’s ability to detect Fatal crashes. The tuned model predicted all observations as the majority Injury class, resulting in Fatal Recall = 0. Therefore, despite slightly better overall metrics, the tuned Naive Bayes model is not suitable for this project because identifying Fatal crashes is a critical objective.
------------------------------------------------------------------------

### 7 Model Evaluation & Selection

<!-- TODO: [member name]  -  comparison table (6 models × 4 metrics), best model selection + justification -->

------------------------------------------------------------------------

### 8 Final Results & Conclusion

<!-- TODO: [member name]  -  answer RQ, top predictors, practical implications, process reflection -->

------------------------------------------------------------------------

### References

<!-- TODO: [member name]  -  APA 7th format references for all sources cited -->

------------------------------------------------------------------------

### Acknowledgement of AI

<!-- TODO: [member name]  -  AI acknowledgement statement -->
