---
title: ""
author: ""
output:
  html_document:
    code_folding: hide
---

```{r library, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(readr)
library(forcats)
library(scales)
library(knitr)
library(caret)
library(nnet)
library(rpart)
library(rpart.plot)
library(randomForest)
library(xgboost)
library(kknn)
library(e1071)
library(themis)
library(pROC)
library(flextable)
library(MLmetrics)
```

```{css, echo=FALSE}
body { font-family: "Times New Roman", Times, serif; font-size: 10pt; line-height: 1.3; }
h2 { margin-top: 0.4em; margin-bottom: 0.15em; }
h3 { margin-top: 0.3em; margin-bottom: 0.1em; }
h4 { margin-top: 0.25em; margin-bottom: 0.1em; }
p  { margin-top: 0.2em; margin-bottom: 0.2em; }
h1.title { display: none; }
h4.author { display: none; }
h4.date   { display: none; }
```

```{r load-data, message=FALSE, warning=FALSE, results='hide'}
df <- read.csv("nsw_road_crash_merged_2020_2024.csv", check.names = FALSE)
cat("Dimensions:", nrow(df), "rows x", ncol(df), "cols\n")
str(df)
```

------------------------------------------------------------------------

### 1 Problem Framing & Research Question

<!-- TODO: [member name] — RQ statement, motivation, 3 stakeholder groups (~0.5 page) -->

------------------------------------------------------------------------

### 2 Dataset & Justification

<!-- TODO: [member name] — data source, merge process, 4-requirement justification table (~0.5 page) -->

------------------------------------------------------------------------

### 3 EDA & Insights

<!-- TODO: [member name] — 4 ggplot2 plots with key finding commentary per plot (~1 page) -->

------------------------------------------------------------------------

### 4 Data Preprocessing

All data-dependent steps are performed **within cross-validation folds** to prevent data leakage (James et al., 2021).

##### Step 1 — Data Integration & TU-Level Feature Aggregation

```{r step1-integration, message=FALSE, warning=FALSE, results='hide'}
# Each row is one traffic unit — summarise binary flags per crash, then join back
tu_flags <- df %>%
  group_by(`Crash ID`) %>%
  summarise(
    has_motorcycle    = as.integer(any(`TU type group` == "Motorcycle")),
    has_heavy_vehicle = as.integer(any(`TU type group` %in% c("Articulated truck", "Heavy rigid truck", "Bus"))),
    has_pedestrian    = as.integer(any(`TU type group` == "Pedestrian")),
    has_cyclist       = as.integer(any(`TU type group` == "Pedal cycle")),
    tu_count          = n(),
    .groups = "drop"
  )

df <- df %>% left_join(tu_flags, by = "Crash ID")
cat("TU flags added. Rows:", nrow(df), "| New cols:", ncol(df), "\n")
```

##### Step 2 — Feature Removal

```{r step2-removal, message=FALSE, warning=FALSE, results='hide'}
# Leakage: post-crash outcome counts that directly encode the target
leakage_cols <- c("No. killed", "No. seriously injured",
                  "No. moderately injured", "No. minor-other injured")

# Redundant: detailed severity is a finer version of the target column
redundant_target_cols <- c("Degree of crash - detailed")

# Identifiers and free-text location fields (no predictive signal)
id_cols <- c("Crash ID", "Traffic unit ID", "Street of crash",
             "Street of travel", "Town", "Identifying feature",
             "Identifying feature type", "Conurbation 1")

# High-NA columns (>75% missing — confirmed from inspection)
high_na_cols <- c("Primary hazardous feature", "Primary temporary feature",
                  "Object hit 2", "DCA supplement", "Object hit 1",
                  "Primary permanent feature")

redundant_cols <- c("Route no.", "RUM - code", "RUM - description",
                    "DCA - code", "DCA - description")

drop_cols <- unique(c(leakage_cols, redundant_target_cols, id_cols, high_na_cols, redundant_cols))
df_clean  <- df %>% select(-any_of(drop_cols))
cat("Columns after removal:", ncol(df_clean), "\n")
```

##### Step 3 — Missing Value Imputation

```{r step3-imputation, message=FALSE, warning=FALSE, results='hide'}
df_clean <- df_clean %>%
  mutate(
    `Other TU type` = ifelse(is.na(`Other TU type`), "Unknown", `Other TU type`),
    across(where(is.character), ~ replace_na(., "Unknown")),
    across(where(is.numeric),   ~ replace_na(., median(., na.rm = TRUE)))
  )
cat("Missing values after imputation:", sum(is.na(df_clean)), "\n")
```

##### Step 4 — Feature Engineering

```{r step4-engineering, message=FALSE, warning=FALSE, results='hide'}
df_clean <- df_clean %>%
  mutate(
    crash_hour   = as.integer(sub(":.*", "", `Two-hour intervals`)),
    is_weekend   = as.integer(`Day of week of crash` %in% c("Saturday", "Sunday")),
    is_rush_hour = as.integer(crash_hour %in% c(7, 8, 9, 16, 17, 18)),
    is_night     = as.integer(`Natural lighting` %in% c("Darkness", "Dawn", "Dusk")),
    is_wet       = as.integer(`Surface condition` %in% c("Wet", "Snow or ice")),
    speed_band   = case_when(
      `Speed limit` %in% c("10 km/h","20 km/h","30 km/h","40 km/h","50 km/h") ~ "low_50",
      `Speed limit` %in% c("60 km/h","70 km/h","80 km/h")                     ~ "mid_60_80",
      `Speed limit` %in% c("90 km/h","100 km/h")                              ~ "high_90_100",
      `Speed limit` == "110 km/h"                                              ~ "very_high_110",
      TRUE                                                                     ~ "unknown"
    )
  )
cat("Engineered features added. Columns:", ncol(df_clean), "\n")
```

##### Step 5 — Outlier Detection

```{r step5-outliers, message=FALSE, warning=FALSE, results='hide'}
p99_dist <- quantile(df_clean$Distance, 0.99, na.rm = TRUE)
df_clean <- df_clean %>%
  mutate(Distance = pmin(Distance, p99_dist)) %>%
  filter(is.na(Latitude)  | (Latitude  >= -37.5 & Latitude  <= -28.2),
         is.na(Longitude) | (Longitude >= 141.0  & Longitude <= 153.6))
cat("Rows after outlier removal:", nrow(df_clean), "\n")
```

##### Step 6 — Encoding & LGA Pooling

```{r step6-encoding, message=FALSE, warning=FALSE, results='hide'}
lga_freq  <- prop.table(table(df_clean$LGA))
rare_lgas <- names(lga_freq[lga_freq < 0.01])
df_clean  <- df_clean %>%
  mutate(LGA = ifelse(LGA %in% rare_lgas, "Other", LGA))

df_model <- df_clean %>%
  mutate(across(where(is.character), as.factor)) %>%
  mutate(`Degree of crash` = factor(`Degree of crash`,
           levels = c("Fatal", "Injury", "Non-casualty (towaway)"),
           labels = c("Fatal", "Injury", "Non_casualty"))) # rename: caret requires valid R variable names

cat("Class distribution:\n")
print(prop.table(table(df_model$`Degree of crash`)))
cat("LGA levels after pooling:", nlevels(df_model$LGA), "\n")
```

##### Step 7 — Train/Test Split & Temporal Holdout

```{r step7-split, message=FALSE, warning=FALSE, results='hide'}
# 2024 sealed as temporal holdout — evaluated once at the very end only
df_holdout_2024  <- df_model %>% filter(`Reporting year` == 2024)
df_model_2020_23 <- df_model %>% filter(`Reporting year` != 2024)

set.seed(42)
train_idx <- createDataPartition(df_model_2020_23$`Degree of crash`, p = 0.8, list = FALSE)
df_train  <- df_model_2020_23[ train_idx, ]
df_test   <- df_model_2020_23[-train_idx, ]

cat("Train:", nrow(df_train), "| Test:", nrow(df_test), "| Holdout:", nrow(df_holdout_2024), "\n")
print(prop.table(table(df_train$`Degree of crash`)))

# df_train_sub = stratified 10% subsample for slow models (RF, XGBoost, KNN).
# df_train     = full training set for fast models (LR, DT, Naive Bayes).
# For final submission: replace df_train_sub with df_train in slow model chunks.
set.seed(42)
sub_idx      <- createDataPartition(df_train$`Degree of crash`, p = 0.50, list = FALSE)
df_train_sub <- df_train[sub_idx, ]
cat("Full train:", nrow(df_train), "rows | Subsample:", nrow(df_train_sub), "rows (10%)\n")
```

##### Step 8 — Class Imbalance Handling

```{r step8-smote, message=FALSE, warning=FALSE, results='hide'}
# Inverse frequency weights for models supporting class weights (LR, DT, RF)
class_weights <- 1 / prop.table(table(df_train$`Degree of crash`))
class_weights <- class_weights / sum(class_weights)
cat("Class weights:\n"); print(round(class_weights, 4))
# SMOTE applied inside CV fold for KNN and Naive Bayes via cv_ctrl_smote
```

##### Step 9 — Cross-Validation Setup

```{r step9-cv, message=FALSE, warning=FALSE, results='hide'}
cv_ctrl <- trainControl(
  method = "cv", number = 5, classProbs = TRUE,
  summaryFunction = multiClassSummary, savePredictions = "final"
)
cv_ctrl_smote <- trainControl(
  method = "cv", number = 5, classProbs = TRUE,
  summaryFunction = multiClassSummary, savePredictions = "final",
  sampling = "smote"
)
cat("CV controls ready.\n")
```

------------------------------------------------------------------------

### 5 Model Fitting

Six models selected to cover a range of complexity and interpretability. All implemented in R via `caret` with 5-fold CV. Primary metric: **Macro-F1**. Hard constraint: **Fatal Recall ≥ 0.50**.

#### 5.1 Multinomial Logistic Regression

```{r model-lr, message=FALSE, warning=FALSE}
# TODO: [member name] — MODEL FITTING ONLY (rubric §5 = basic fit, tuning goes in §6.1)
# Use DEFAULT hyperparameters — do NOT tune here
# method = "multinom" | preProcess = "nzv" | CV: cv_ctrl_smote | Weights: class_weights
# Produce: confusionMatrix() on df_test, heatmap, Macro-F1 / Fatal Recall / Weighted F1 / Kappa
# Save metrics as: macro_f1_lr, fatal_recall_lr (needed for §6.1 before/after table)

```

#### 5.2 Decision Tree

```{r model-dt, message=FALSE, warning=FALSE}
# TODO: [member name] — MODEL FITTING ONLY (rubric §5 = basic fit, tuning goes in §6.2)
# Use DEFAULT hyperparameters — do NOT tune here
# method = "rpart" | CV: cv_ctrl | Weights: class_weights (via weights argument)
# Produce: rpart.plot diagram, confusionMatrix() on df_test, heatmap, 4 metrics
# Save metrics as: macro_f1_dt, fatal_recall_dt (needed for §6.2 before/after table)

```

#### 5.3 Random Forest

```{r model-rf, message=FALSE, warning=FALSE, fig.show='hold', out.width='50%'}
# Call randomForest() directly — caret's train() CV wrapper fails on this dataset:
# Fatal class (~1.4%) causes some CV folds to produce zero Fatal predictions,
# making all fold metrics NA and crashing caret. RF's built-in OOB error is an
# unbiased estimate of generalisation error and makes CV redundant anyway.

# randomForest cannot handle factors with >53 levels or NAs in predictors.
# na.roughfix() imputes residual NAs: numeric → column median, factor → mode.
# Using df_train_sub (10%) — swap to df_train for final submission
rf_x_train <- na.roughfix(df_train_sub %>%
  select(-`Degree of crash`) %>%
  select(where(~ !is.factor(.) || nlevels(.) <= 53)) %>%
  droplevels())

# Align test factor levels to training — unseen levels become NA, then imputed
rf_x_test_raw <- df_test %>% select(all_of(names(rf_x_train)))
for (col in names(rf_x_test_raw)) {
  if (is.factor(rf_x_test_raw[[col]]))
    rf_x_test_raw[[col]] <- factor(rf_x_test_raw[[col]], levels = levels(rf_x_train[[col]]))
}
rf_x_test <- na.roughfix(rf_x_test_raw)

set.seed(42)
model_rf <- randomForest(
  x          = rf_x_train,
  y          = df_train_sub$`Degree of crash`,
  mtry       = floor(sqrt(ncol(rf_x_train))),
  ntree      = 100,
  importance = TRUE
)

# Test set predictions and metrics
pred_rf         <- predict(model_rf, rf_x_test)
cm_rf           <- confusionMatrix(pred_rf, df_test$`Degree of crash`)
fatal_recall_rf <- cm_rf$byClass["Class: Fatal", "Recall"]
macro_f1_rf     <- mean(cm_rf$byClass[, "F1"],  na.rm = TRUE)
weighted_f1_rf  <- weighted.mean(cm_rf$byClass[, "F1"],
                    prop.table(table(df_test$`Degree of crash`)), na.rm = TRUE)
kappa_rf        <- cm_rf$overall["Kappa"]

cat(sprintf(
  "=== Random Forest — Test Set Metrics ===\nOOB error rate : %.4f\nMacro-F1       : %.4f\nFatal Recall   : %.4f\nWeighted F1    : %.4f\nKappa          : %.4f\n",
  model_rf$err.rate[100, "OOB"], macro_f1_rf, fatal_recall_rf, weighted_f1_rf, kappa_rf
))

# Confusion matrix heatmap
as.data.frame(cm_rf$table) %>%
  ggplot(aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = Freq), size = 6, fontface = "bold") +
  scale_fill_gradient(low = "white", high = "#1F4E79") +
  labs(title = "Random Forest — Confusion Matrix (Test Set)",
       subtitle = paste0("Fatal Recall: ", round(fatal_recall_rf, 3),
                         "  |  Macro-F1: ", round(macro_f1_rf, 3))) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")

# Variable importance (top 15) — MeanDecreaseAccuracy across all classes
imp_df <- as.data.frame(importance(model_rf, type = 1))
imp_df$Feature <- rownames(imp_df)
imp_df %>%
  arrange(desc(MeanDecreaseAccuracy)) %>%
  head(15) %>%
  ggplot(aes(x = reorder(Feature, MeanDecreaseAccuracy), y = MeanDecreaseAccuracy)) +
  geom_col(fill = "#2E74B5") +
  coord_flip() +
  labs(title = "Random Forest — Variable Importance (Top 15)",
       x = NULL, y = "Mean Decrease in Accuracy") +
  theme_minimal(base_size = 12)
```

#### 5.4 XGBoost

```{r model-xgb, message=FALSE, warning=FALSE}
# TODO: [member name] — MODEL FITTING ONLY (rubric §5 = basic fit, tuning goes in §6.4)
# Use DEFAULT hyperparameters — do NOT tune here
# method = "xgbTree" | requires one-hot encoding via model.matrix(~ . - 1)
# CV: cv_ctrl | Weights: scale_pos_weight in xgb params
# Produce: confusionMatrix() on df_test, heatmap, 4 metrics
# Save metrics as: macro_f1_xgb, fatal_recall_xgb (needed for §6.4 before/after table)

```

#### 5.5 K-Nearest Neighbours

```{r model-knn, message=FALSE, warning=FALSE}
# TODO: [member name] — MODEL FITTING ONLY (rubric §5 = basic fit, tuning goes in §6.5)
# Use DEFAULT k — do NOT tune here (default k = 5 or caret's single default)
# method = "kknn" | preProcess = c("center", "scale") | CV: cv_ctrl_smote
# Produce: confusionMatrix() on df_test, heatmap, 4 metrics
# Save metrics as: macro_f1_knn, fatal_recall_knn (needed for §6.5 before/after table)

```

#### 5.6 Naive Bayes

```{r model-nb, message=FALSE, warning=FALSE}
# TODO: [member name] — MODEL FITTING ONLY (rubric §5 = basic fit, tuning goes in §6.6)
# Use DEFAULT hyperparameters (usekernel = FALSE, laplace = 0) — do NOT tune here
# method = "naive_bayes" | CV: cv_ctrl_smote | Weights: via SMOTE + class prior
# Produce: confusionMatrix() on df_test, heatmap, 4 metrics
# Save metrics as: macro_f1_nb, fatal_recall_nb (needed for §6.6 before/after table)

```

------------------------------------------------------------------------

### 6 Model Improvement

#### 6.1 Multinomial Logistic Regression — Improvement

```{r improve-lr, message=FALSE, warning=FALSE}
# TODO: [member name] — MODEL IMPROVEMENT (rubric §6 = tuning/improvement, basic fit is in §5.1)
# Tune: decay via tuneGrid = expand.grid(decay = c(0, 0.001, 0.01, 0.1))
# Plot: decay vs Mean_F1 curve from model_lr_tuned$results
# Table: before/after using macro_f1_lr / fatal_recall_lr from §5.1 vs tuned values
```

#### 6.2 Decision Tree — Improvement

```{r improve-dt, message=FALSE, warning=FALSE}
# TODO: [member name] — MODEL IMPROVEMENT (rubric §6 = tuning/improvement, basic fit is in §5.2)
# Tune: cp via tuneGrid = expand.grid(cp = seq(0.0001, 0.01, by = 0.001))
# Plot: cp vs Mean_F1 curve from model_dt_tuned$results (or plotcp on standalone rpart fit)
# Table: before/after using macro_f1_dt / fatal_recall_dt from §5.2 vs tuned values
```

#### 6.3 Random Forest — Improvement

```{r improve-rf, message=FALSE, warning=FALSE, fig.show='hold', out.width='50%'}
# Use a 20k-row subsample of rf_x_train to search mtry quickly via tuneRF().
# Finding the best mtry doesn't need the full dataset — OOB rankings are stable
# on a representative subsample. Final model is still fit on the full training set.
set.seed(42)
tune_idx <- sample(nrow(rf_x_train), min(nrow(rf_x_train), 20000))
invisible(capture.output(
  tune_result <- tuneRF(
    x          = rf_x_train[tune_idx, ],
    y          = df_train_sub$`Degree of crash`[tune_idx],
    mtryStart  = floor(sqrt(ncol(rf_x_train))),
    ntreeTry   = 50,
    stepFactor = 1.5,
    improve    = 0.01,
    trace      = FALSE,
    plot       = FALSE
  )
))
best_mtry <- tune_result[which.min(tune_result[, "OOBError"]), "mtry"]
cat("Best mtry from tuneRF:", best_mtry, "\n")

# mtry vs OOB error curve
as.data.frame(tune_result) %>%
  ggplot(aes(x = mtry, y = OOBError)) +
  geom_line(colour = "#2E74B5", linewidth = 1) +
  geom_point(size = 3, colour = "#2E74B5") +
  geom_vline(xintercept = best_mtry, linetype = "dashed", colour = "#C0392B") +
  annotate("text", x = best_mtry, y = max(tune_result[, "OOBError"]),
           label = paste("Best mtry =", best_mtry),
           hjust = -0.1, colour = "#C0392B", size = 3.5) +
  labs(title = "Random Forest — Tuning: mtry vs OOB Error (20k subsample)",
       x = "mtry (variables randomly sampled per split)", y = "OOB Error Rate") +
  theme_minimal(base_size = 12)

# Final tuned model fit — using df_train_sub (50%), swap to df_train for final submission
set.seed(42)
model_rf_tuned <- randomForest(
  x          = rf_x_train,
  y          = df_train_sub$`Degree of crash`,
  mtry       = best_mtry,
  ntree      = 100,
  importance = TRUE
)

pred_rf_tuned      <- predict(model_rf_tuned, rf_x_test)
cm_rf_tuned        <- confusionMatrix(pred_rf_tuned, df_test$`Degree of crash`)
fatal_recall_tuned <- cm_rf_tuned$byClass["Class: Fatal", "Recall"]
macro_f1_tuned     <- mean(cm_rf_tuned$byClass[, "F1"], na.rm = TRUE)
weighted_f1_tuned  <- weighted.mean(cm_rf_tuned$byClass[, "F1"],
                       prop.table(table(df_test$`Degree of crash`)), na.rm = TRUE)
kappa_tuned        <- cm_rf_tuned$overall["Kappa"]

cat(sprintf(
  "=== RF Tuned — Test Set Metrics ===\nOOB error rate : %.4f\nMacro-F1       : %.4f\nFatal Recall   : %.4f\nWeighted F1    : %.4f\nKappa          : %.4f\n",
  model_rf_tuned$err.rate[100, "OOB"], macro_f1_tuned, fatal_recall_tuned, weighted_f1_tuned, kappa_tuned
))

# Before / after comparison table
data.frame(
  Model          = c("RF — default mtry", "RF — tuned mtry"),
  mtry           = c(floor(sqrt(ncol(rf_x_train))), best_mtry),
  `Macro-F1`     = round(c(macro_f1_rf,    macro_f1_tuned),    4),
  `Fatal Recall` = round(c(fatal_recall_rf, fatal_recall_tuned), 4),
  check.names = FALSE
) %>% knitr::kable(caption = "Random Forest — Before vs After Tuning")

# OOB error vs ntree curve (convergence check)
data.frame(ntree = seq_len(nrow(model_rf_tuned$err.rate)),
           OOB   = model_rf_tuned$err.rate[, "OOB"]) %>%
  ggplot(aes(x = ntree, y = OOB)) +
  geom_line(colour = "#2E74B5") +
  labs(title = "Random Forest — OOB Error vs Number of Trees (tuned mtry)",
       x = "Number of trees", y = "OOB error rate") +
  theme_minimal(base_size = 12)
```

#### 6.4 XGBoost — Improvement

```{r improve-xgb, message=FALSE, warning=FALSE}
# TODO: [member name] — MODEL IMPROVEMENT (rubric §6 = tuning/improvement, basic fit is in §5.4)
# Tune: nrounds, max_depth, eta via expand.grid(nrounds=c(50,100,200), max_depth=c(3,6), eta=c(0.01,0.1,0.3))
# Plot: best nrounds vs Mean_F1 curve from model_xgb_tuned$results
# Table: before/after using macro_f1_xgb / fatal_recall_xgb from §5.4 vs tuned values
```

#### 6.5 K-Nearest Neighbours — Improvement

```{r improve-knn, message=FALSE, warning=FALSE}
# TODO: [member name] — MODEL IMPROVEMENT (rubric §6 = tuning/improvement, basic fit is in §5.5)
# Tune: kmax via tuneGrid = expand.grid(kmax = c(5, 11, 21, 31, 51), distance = 2, kernel = "optimal")
# Plot: k vs Mean_F1 curve from model_knn_tuned$results
# Table: before/after using macro_f1_knn / fatal_recall_knn from §5.5 vs tuned values
```

#### 6.6 Naive Bayes — Improvement

```{r improve-nb, message=FALSE, warning=FALSE}
# TODO: [member name] — MODEL IMPROVEMENT (rubric §6 = tuning/improvement, basic fit is in §5.6)
# Tune: usekernel and laplace via tuneGrid = expand.grid(usekernel = c(TRUE, FALSE), laplace = c(0, 0.5, 1), adjust = 1)
# Plot: bar chart of Macro-F1 across tuning combinations
# Table: before/after using macro_f1_nb / fatal_recall_nb from §5.6 vs tuned values
```

------------------------------------------------------------------------

### 7 Model Evaluation & Selection

<!-- TODO: [member name] — comparison table (6 models × 4 metrics), best model selection + justification -->

------------------------------------------------------------------------

### 8 Final Results & Conclusion

<!-- TODO: [member name] — answer RQ, top predictors, practical implications, process reflection -->

------------------------------------------------------------------------

### References

<!-- TODO: [member name] — APA 7th format references for all sources cited -->

### Acknowledgement of AI

<!-- TODO: [member name] — AI acknowledgement statement -->
