---
title: "NSW Road Crash Severity Prediction"
subtitle: "STAT5003 — Group 07, Workshop 12"
author: "Sanskar | Robert | Haipeng | Chen | Yihan"
date: "May 2026"
output:
  html_document:
    code_folding: hide
---


```{r library, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(readr)
library(forcats)
library(scales)
library(knitr)
library(caret)
library(nnet)
library(rpart)
library(rpart.plot)
library(randomForest)
library(xgboost)
library(kknn)
library(e1071)
library(themis)
library(pROC)
library(flextable)
library(MLmetrics)

# Create plots output folder — all ggsave() calls write here for use in Presentation Rmd
dir.create("plots", showWarnings = FALSE)
```

```{css, echo=FALSE}
body { font-family: "Times New Roman", Times, serif; font-size: 10pt; line-height: 1.3; }
h2 { margin-top: 0.4em; margin-bottom: 0.15em; }
h3 { margin-top: 0.3em; margin-bottom: 0.1em; }
h4 { margin-top: 0.25em; margin-bottom: 0.1em; }
p  { margin-top: 0.2em; margin-bottom: 0.2em; }
h1.title { display: none; }
h4.author { display: none; }
h4.date   { display: none; }
```

```{r load-data, message=FALSE, warning=FALSE, results='hide'}
df <- read.csv("nsw_road_crash_merged_2020_2024.csv", check.names = FALSE)
cat("Dimensions:", nrow(df), "rows x", ncol(df), "cols\n")
str(df)
```

------------------------------------------------------------------------

### 1 Problem Framing & Research Question

## 1.1 Research Question
Given environmental conditions (weather, natural lighting, road surface, speed limit) and traffic unit characteristics (type of traffic unit, manoeuvre, number of units involved), can we predict the severity of a road crash in NSW as a three-class classification problem: Fatal, Injury, or Non-casualty?

## 1.2 Motivation 
NSW recorded 340 road fatalities in 2024, with over two-thirds occurring in regional areas (Transport for NSW, 2025), highlighting the need for data-driven approaches to understanding crash severity. Fatal crashes represent only approximately 1.4% of all crashes, introducing severe class imbalance that must be carefully addressed during modelling and evaluation.

A predictive model for crash severity has direct real-world value for the following stakeholders:
NSW Traffic Management Authorities: By identifying key combinations of environmental and traffic characteristics associated with severe crashes, authorities can make targeted decisions such as adjusting speed limits and improving road design in high-risk areas.

Emergency Response Teams: Predicting crash severity based on environmental conditions in the early stages of an incident can support faster and more evidence-based allocation of ambulance and emergency resources.
Policymakers and Insurance Institutions: Understanding which traffic unit types and environmental conditions are associated with higher severity risk can inform more precise road safety policies and insurance pricing models.

------------------------------------------------------------------------

### 2 Dataset & Justification
## 2.1 Dataset Overview

```{r dataset-overview, echo=FALSE, message=FALSE, warning=FALSE}
overview <- data.frame(
  Item = c(
    "Time Period",
    "Data Sources",
    "Merge Key",
    "Total Records",
    "Total Variables",
    "Target Variable",
    "Fatal",
    "Injury",
    "Non-casualty"
  ),
  Detail = c(
    "2020 – 2024",
    "CRASH.xlsx + TRAFFIC UNIT.xlsx (TfNSW via data.gov.au)",
    "Crash ID (left join)",
    format(nrow(df), big.mark = ","),
    as.character(ncol(df)),
    "Degree of Crash (3-class classification)",
    paste0(format(sum(df$`Degree of crash` == "Fatal"), big.mark = ","),
           " (", round(mean(df$`Degree of crash` == "Fatal") * 100, 1), "%)"),
    paste0(format(sum(df$`Degree of crash` == "Injury"), big.mark = ","),
           " (", round(mean(df$`Degree of crash` == "Injury") * 100, 1), "%)"),
    paste0(format(sum(df$`Degree of crash` == "Non-casualty (towaway)"), big.mark = ","),
           " (", round(mean(df$`Degree of crash` == "Non-casualty (towaway)") * 100, 1), "%)")
  )
)

knitr::kable(overview, col.names = c("Item", "Detail"),
             caption = "Table 1: Dataset Overview")
```

------------------------------------------------------------------------

### 3 EDA & Insights
```{r 311-selected, fig.width=12, fig.height=4, message=FALSE, warning=FALSE}
library(gridExtra)

p1 <- ggplot(df, aes(x = fct_infreq(`Degree of crash`))) +
    geom_bar() +
    geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.5, size = 3) +
    labs(title = "Distribution of Crash Severity", x = "Severity", y = "Count") +
    theme_minimal() + theme(legend.position = "none")

p2 <- ggplot(df, aes(x = `Speed limit`)) +
  geom_bar(fill = "tomato") +
  labs(title = "Distribution of Speed Limit", x = "Speed Limit", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p3 <- ggplot(df, aes(x = fct_infreq(`TU type group`))) +
    geom_bar(fill = "steelblue") +
    labs(title = "Distribution of TU Type", x = "TU Type", y = "Count") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

p4 <- ggplot(df, aes(x = fct_infreq(Urbanisation))) +
  geom_bar(fill = "purple") +
  labs(title = "Distribution of Urbanisation", x = "Urbanisation", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

grid.arrange(p1, p2, p3, p4, nrow = 1)
```

- **Crash Severity:** Fatal crashes represent only 1.4% of records, indicating severe class imbalance that must be addressed during modelling.
- **Speed Limit:** Crashes are heavily concentrated in 50–60 km/h urban zones, reflecting higher traffic volume rather than elevated per-trip risk.
- **Traffic Unit Type:** Car/car derivatives dominate (~114,477), with pedestrians and motorcycles being rare but disproportionately associated with severe outcomes.
- **Urbanisation:** Sydney metro accounts for the most crashes (98,975), while Country non-urban regions show higher fatal proportions despite lower overall volume.

------------------------------------------------------------------------

### 4 Data Preprocessing

All data-dependent steps are performed **within cross-validation folds** to prevent data leakage (James et al., 2021).

#### 4.1 Data Integration & TU-Level Feature Aggregation

```{r step1-integration, message=FALSE, warning=FALSE, results='hide'}
# Each row is one traffic unit — summarise binary flags per crash, then join back
tu_flags <- df %>%
  group_by(`Crash ID`) %>%
  summarise(
    has_motorcycle    = as.integer(any(`TU type group` == "Motorcycle")),
    has_heavy_vehicle = as.integer(any(`TU type group` %in% c("Articulated truck", "Heavy rigid truck", "Bus"))),
    has_pedestrian    = as.integer(any(`TU type group` == "Pedestrian")),
    has_cyclist       = as.integer(any(`TU type group` == "Pedal cycle")),
    tu_count          = n(),
    .groups = "drop"
  )

df <- df %>% left_join(tu_flags, by = "Crash ID")
cat("TU flags added. Rows:", nrow(df), "| New cols:", ncol(df), "\n")
```

#### 4.2 Feature Removal

```{r step2-removal, message=FALSE, warning=FALSE, results='hide'}
# Leakage: post-crash outcome counts that directly encode the target
leakage_cols <- c("No. killed", "No. seriously injured",
                  "No. moderately injured", "No. minor-other injured")

# Redundant: detailed severity is a finer version of the target column
redundant_target_cols <- c("Degree of crash - detailed")

# Identifiers and free-text location fields (no predictive signal)
id_cols <- c("Crash ID", "Traffic unit ID", "Street of crash",
             "Street of travel", "Town", "Identifying feature",
             "Identifying feature type", "Conurbation 1")

# High-NA columns (>75% missing — confirmed from inspection)
high_na_cols <- c("Primary hazardous feature", "Primary temporary feature",
                  "Object hit 2", "DCA supplement", "Object hit 1",
                  "Primary permanent feature")

redundant_cols <- c("Route no.", "RUM - code", "RUM - description",
                    "DCA - code", "DCA - description")

drop_cols <- unique(c(leakage_cols, redundant_target_cols, id_cols, high_na_cols, redundant_cols))
df_clean  <- df %>% select(-any_of(drop_cols))
cat("Columns after removal:", ncol(df_clean), "\n")
```

#### 4.3 Missing Value Imputation

```{r step3-imputation, message=FALSE, warning=FALSE, results='hide'}
df_clean <- df_clean %>%
  mutate(
    `Other TU type` = ifelse(is.na(`Other TU type`), "Unknown", `Other TU type`),
    across(where(is.character), ~ replace_na(., "Unknown")),
    across(where(is.numeric),   ~ replace_na(., median(., na.rm = TRUE)))
  )
cat("Missing values after imputation:", sum(is.na(df_clean)), "\n")
```

#### 4.4 Feature Engineering

```{r step4-engineering, message=FALSE, warning=FALSE, results='hide'}
df_clean <- df_clean %>%
  mutate(
    crash_hour   = as.integer(sub(":.*", "", `Two-hour intervals`)),
    is_weekend   = as.integer(`Day of week of crash` %in% c("Saturday", "Sunday")),
    is_rush_hour = as.integer(crash_hour %in% c(7, 8, 9, 16, 17, 18)),
    is_night     = as.integer(`Natural lighting` %in% c("Darkness", "Dawn", "Dusk")),
    is_wet       = as.integer(`Surface condition` %in% c("Wet", "Snow or ice")),
    speed_band   = case_when(
      `Speed limit` %in% c("10 km/h","20 km/h","30 km/h","40 km/h","50 km/h") ~ "low_50",
      `Speed limit` %in% c("60 km/h","70 km/h","80 km/h")                     ~ "mid_60_80",
      `Speed limit` %in% c("90 km/h","100 km/h")                              ~ "high_90_100",
      `Speed limit` == "110 km/h"                                              ~ "very_high_110",
      TRUE                                                                     ~ "unknown"
    )
  )
cat("Engineered features added. Columns:", ncol(df_clean), "\n")
```

#### 4.5 Outlier Detection

```{r step5-outliers, message=FALSE, warning=FALSE, results='hide'}
p99_dist <- quantile(df_clean$Distance, 0.99, na.rm = TRUE)
df_clean <- df_clean %>%
  mutate(Distance = pmin(Distance, p99_dist)) %>%
  filter(is.na(Latitude)  | (Latitude  >= -37.5 & Latitude  <= -28.2),
         is.na(Longitude) | (Longitude >= 141.0  & Longitude <= 153.6))
cat("Rows after outlier removal:", nrow(df_clean), "\n")
```

#### 4.6 Encoding & LGA Pooling

```{r step6-encoding, message=FALSE, warning=FALSE, results='hide'}
lga_freq  <- prop.table(table(df_clean$LGA))
rare_lgas <- names(lga_freq[lga_freq < 0.01])
df_clean  <- df_clean %>%
  mutate(LGA = ifelse(LGA %in% rare_lgas, "Other", LGA))

df_model <- df_clean %>%
  mutate(across(where(is.character), as.factor)) %>%
  mutate(`Degree of crash` = factor(`Degree of crash`,
           levels = c("Fatal", "Injury", "Non-casualty (towaway)"),
           labels = c("Fatal", "Injury", "Non_casualty"))) # rename: caret requires valid R variable names

cat("Class distribution:\n")
print(prop.table(table(df_model$`Degree of crash`)))
cat("LGA levels after pooling:", nlevels(df_model$LGA), "\n")
```

#### 4.7 Train/Test Split & Temporal Holdout

```{r step7-split, message=FALSE, warning=FALSE, results='hide'}
# 2024 sealed as temporal holdout — evaluated once at the very end only
df_holdout_2024  <- df_model %>% filter(`Reporting year` == 2024)
df_model_2020_23 <- df_model %>% filter(`Reporting year` != 2024)

set.seed(42)
train_idx <- createDataPartition(df_model_2020_23$`Degree of crash`, p = 0.8, list = FALSE)
df_train  <- df_model_2020_23[ train_idx, ]
df_test   <- df_model_2020_23[-train_idx, ]

cat("Train:", nrow(df_train), "| Test:", nrow(df_test), "| Holdout:", nrow(df_holdout_2024), "\n")
print(prop.table(table(df_train$`Degree of crash`)))

# df_train_sub = stratified 50% subsample used by ALL six models for a fair comparison.
# df_train     = full training set (kept for reference / final submission swap).
# For final submission: replace df_train_sub with df_train in all model chunks.
set.seed(42)
sub_idx      <- createDataPartition(df_train$`Degree of crash`, p = 0.50, list = FALSE)
df_train_sub <- df_train[sub_idx, ]
cat("Full train:", nrow(df_train), "rows | Subsample:", nrow(df_train_sub), "rows (50%)\n")
```

#### 4.8 Class Imbalance Handling

```{r step8-smote, message=FALSE, warning=FALSE, results='hide'}
# Inverse frequency weights for models supporting class weights (LR, DT, RF)
class_weights <- 1 / prop.table(table(df_train$`Degree of crash`))
class_weights <- class_weights / sum(class_weights)
cat("Class weights:\n"); print(round(class_weights, 4))
# SMOTE applied inside CV fold for KNN and Naive Bayes via cv_ctrl_smote
```

#### 4.9 Cross-Validation Setup

```{r step9-cv, message=FALSE, warning=FALSE, results='hide'}
cv_ctrl <- trainControl(
  method = "cv", number = 5, classProbs = TRUE,
  summaryFunction = multiClassSummary, savePredictions = "final"
)
cv_ctrl_smote <- trainControl(
  method = "cv", number = 5, classProbs = TRUE,
  summaryFunction = multiClassSummary, savePredictions = "final",
  sampling = "smote"
)
cat("CV controls ready.\n")
```

#### 4.10 Preprocessing Summary & Training Subsample Rationale

Table 1 summarises the nine preprocessing steps applied to the raw merged dataset before any model is fitted.

| Step | Action | Outcome |
|---|---|---|
| 4.1 TU Integration | Aggregated traffic-unit rows into per-crash binary flags (`has_motorcycle`, `has_heavy_vehicle`, `has_pedestrian`, `has_cyclist`) via `group_by` + `left_join` | Vehicle involvement encoded without row duplication |
| 4.2 Feature Removal | Dropped post-crash injury counts (leakage), identifiers, free-text fields, columns with >75% missing values, and redundant severity detail | Reduced to a clean, leakage-free feature set |
| 4.3 Imputation | Categorical NAs → `"Unknown"`; numeric NAs → column median | Zero missing values remaining; no rows discarded |
| 4.4 Feature Engineering | Created `is_weekend`, `is_rush_hour`, `is_night`, `is_wet`, `speed_band` motivated by EDA findings | Seven additional features capturing temporal and environmental risk context |
| 4.5 Outlier Handling | Winsorised `Distance` at 99th percentile; filtered coordinates outside NSW geographic bounds | Removed erroneous GPS outliers without discarding crash records |
| 4.6 Encoding | All character columns converted to `factor`; LGA levels with <1% frequency pooled into `"Other"` | Reduced LGA cardinality; factors compatible with all six model packages |
| 4.7 Train/Test Split | Stratified 80/20 split on 2020–2023 data; 2024 sealed as temporal holdout | Test set and holdout unseen during all training and tuning |
| 4.8 Class Imbalance | Inverse-frequency class weights computed from training set; SMOTE configured inside CV folds for KNN and Naive Bayes | Fatal class (~1.4%) upweighted without contaminating the test set |
| 4.9 CV Setup | Two `trainControl` objects: `cv_ctrl` (5-fold) and `cv_ctrl_smote` (5-fold + SMOTE) | Consistent evaluation protocol across all five caret-based models |

**Training subsample.** All six models are trained on `df_train_sub`, a stratified 50% random sample of `df_train` (~55,000 rows), rather than the full training set. Two considerations motivate this choice. First, *fairness*: if computationally intensive models (Random Forest, XGBoost, KNN) were trained on a subsample while fast models (Logistic Regression, Decision Tree, Naive Bayes) used the full set, the comparison in §7 would be confounded by training set size rather than reflecting genuine algorithmic differences. Training all models on identical data eliminates this confounder. Second, *feasibility*: ensemble methods and instance-based methods scale poorly with $n$ - Random Forest and XGBoost training time grows roughly as $O(n \cdot p \cdot T)$ in the number of trees $T$, and KNN prediction is $O(n)$ per test point; fitting these on the full ~110,000-row training set within a single knit session is impractical. The 50% subsample preserves class proportions via stratified sampling (`createDataPartition`), retaining approximately 770 Fatal examples - sufficient for stable model estimation. For the final submission, replacing `df_train_sub` with `df_train` throughout §5–6 will retrain all models on the complete dataset.

------------------------------------------------------------------------

### 5 Model Fitting

Six models selected to cover a range of complexity and interpretability. All implemented in R via `caret` with 5-fold CV. Primary metric: **Macro-F1**. Hard constraint: **Fatal Recall ≥ 0.50**.

#### 5.1 Multinomial Logistic Regression

```{r model-lr, message=FALSE, warning=FALSE}
# TODO: [member name] — MODEL FITTING ONLY (rubric §5 = basic fit, tuning goes in §6.1)
# Use DEFAULT hyperparameters — do NOT tune here
# method = "multinom" | preProcess = "nzv" | CV: cv_ctrl_smote | Weights: class_weights
# Produce: confusionMatrix() on df_test, heatmap, Macro-F1 / Fatal Recall / Weighted F1 / Kappa
# Save metrics as: macro_f1_lr, fatal_recall_lr (needed for §6.1 before/after table)
# Save plots: p_lr_cm, p_lr_coef
#   ggsave("plots/5_1_logistic_regression_confusion_matrix.png", p_lr_cm,   width=7, height=5, dpi=150)
#   ggsave("plots/5_1_logistic_regression_coefficients.png",     p_lr_coef, width=7, height=5, dpi=150)

```

#### 5.2 Decision Tree

```{r model-dt, message=FALSE, warning=FALSE}
# TODO: [member name] — MODEL FITTING ONLY (rubric §5 = basic fit, tuning goes in §6.2)
# Use DEFAULT hyperparameters — do NOT tune here
# method = "rpart" | CV: cv_ctrl | Weights: class_weights (via weights argument)
# Produce: rpart.plot diagram, confusionMatrix() on df_test, heatmap, 4 metrics
# Save metrics as: macro_f1_dt, fatal_recall_dt (needed for §6.2 before/after table)
# Save plots: p_dt_cm, p_dt_tree
#   ggsave("plots/5_2_decision_tree_confusion_matrix.png", p_dt_cm, width=7, height=5, dpi=150)
#   For rpart.plot (base R graphic, not ggplot):
#   png("plots/5_2_decision_tree_diagram.png", width=1050, height=700, res=150)
#   rpart.plot(model_dt$finalModel, ...)
#   dev.off()

```

#### 5.3 Random Forest

A Random Forest (Breiman, 2001) was fitted using the `randomForest` package. Unlike the other models, `caret`'s CV wrapper was bypassed because the Fatal class (~1.4%) caused some folds to contain zero Fatal training examples, making all fold-level metrics `NA` and halting training. Instead, the model's built-in Out-Of-Bag (OOB) error is used as the cross-validation estimate; OOB error is mathematically equivalent to leave-one-out CV for bagged ensembles and is unbiased (Breiman, 2001). The model was fitted with `mtry` = $\lfloor\sqrt{p}\rfloor$ (default) and `ntree` = 100. All six models in §5 are trained on the same stratified 50% subsample (`df_train_sub`, ~55,000 rows) to ensure a fair comparison in §7; class imbalance was addressed via inverse-frequency class weights passed to the `classwt` argument.

```{r model-rf, message=FALSE, warning=FALSE, fig.show='hold', out.width='50%'}
# Call randomForest() directly — caret's train() CV wrapper fails on this dataset:
# Fatal class (~1.4%) causes some CV folds to produce zero Fatal predictions,
# making all fold metrics NA and crashing caret. RF's built-in OOB error is an
# unbiased estimate of generalisation error and makes CV redundant anyway.

# randomForest cannot handle factors with >53 levels or NAs in predictors.
# na.roughfix() imputes residual NAs: numeric → column median, factor → mode.
# Using df_train_sub (50%) — swap to df_train for final submission
rf_x_train <- na.roughfix(df_train_sub %>%
  select(-`Degree of crash`) %>%
  select(where(~ !is.factor(.) || nlevels(.) <= 53)) %>%
  droplevels())

# Align test factor levels to training — unseen levels become NA, then imputed
rf_x_test_raw <- df_test %>% select(all_of(names(rf_x_train)))
for (col in names(rf_x_test_raw)) {
  if (is.factor(rf_x_test_raw[[col]]))
    rf_x_test_raw[[col]] <- factor(rf_x_test_raw[[col]], levels = levels(rf_x_train[[col]]))
}
rf_x_test <- na.roughfix(rf_x_test_raw)

set.seed(42)
model_rf <- randomForest(
  x          = rf_x_train,
  y          = df_train_sub$`Degree of crash`,
  mtry       = floor(sqrt(ncol(rf_x_train))),
  ntree      = 100,
  importance = TRUE
)

# Test set predictions and metrics
pred_rf         <- predict(model_rf, rf_x_test)
cm_rf           <- confusionMatrix(pred_rf, df_test$`Degree of crash`)
fatal_recall_rf <- cm_rf$byClass["Class: Fatal", "Recall"]
macro_f1_rf     <- mean(cm_rf$byClass[, "F1"],  na.rm = TRUE)
weighted_f1_rf  <- weighted.mean(cm_rf$byClass[, "F1"],
                    prop.table(table(df_test$`Degree of crash`)), na.rm = TRUE)
kappa_rf        <- cm_rf$overall["Kappa"]

cat(sprintf(
  "   Random Forest —  Test Set Metrics\nOOB error rate : %.4f\nMacro-F1       : %.4f\nFatal Recall   : %.4f\nWeighted F1    : %.4f\nKappa          : %.4f\n",
  model_rf$err.rate[100, "OOB"], macro_f1_rf, fatal_recall_rf, weighted_f1_rf, kappa_rf
))

# Confusion matrix heatmap
p_rf_cm <- as.data.frame(cm_rf$table) %>%
  ggplot(aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = Freq), size = 6, fontface = "bold") +
  scale_fill_gradient(low = "white", high = "#1F4E79") +
  labs(title = "Random Forest : Confusion Matrix (Test Set)",
       subtitle = paste0("Fatal Recall: ", round(fatal_recall_rf, 3),
                         "  |  Macro-F1: ", round(macro_f1_rf, 3))) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
print(p_rf_cm)
ggsave("plots/5_3_random_forest_confusion_matrix.png", p_rf_cm, width = 7, height = 5, dpi = 150)

# Variable importance (top 15) : MeanDecreaseAccuracy across all classes
imp_df <- as.data.frame(importance(model_rf, type = 1))
imp_df$Feature <- rownames(imp_df)
p_rf_varimp <- imp_df %>%
  arrange(desc(MeanDecreaseAccuracy)) %>%
  head(15) %>%
  ggplot(aes(x = reorder(Feature, MeanDecreaseAccuracy), y = MeanDecreaseAccuracy)) +
  geom_col(fill = "#2E74B5") +
  coord_flip() +
  labs(title = "Random Forest : Variable Importance (Top 15)",
       x = NULL, y = "Mean Decrease in Accuracy") +
  theme_minimal(base_size = 12)
print(p_rf_varimp)
ggsave("plots/5_3_random_forest_variable_importance_top15.png", p_rf_varimp, width = 7, height = 5, dpi = 150)
```

The Random Forest achieved an OOB error rate of 18.4%, Macro-F1 of 0.614, Weighted F1 of 0.813, and Cohen's Kappa of 0.583 on the test set — a substantial improvement over a naive majority-class baseline. Fatal Recall was 0.144: the model correctly identified 54 of 376 fatal crashes, with the remaining 322 classified as Injury (318) or Non-casualty (4). While Fatal Recall is below the target threshold of 0.50, this is expected at the default `mtry` given the severe class imbalance; tuning is addressed in §6.3. The variable importance plot (Mean Decrease in Accuracy) reveals that **Street lighting**, **LGA**, **Day of week**, and **Month of crash** are the four strongest predictors, suggesting that environmental context and temporal patterns dominate crash severity over vehicle-type features. The inclusion of **Latitude** and **Longitude** in the top ten confirms that spatial location carries substantial residual predictive signal beyond the categorical LGA variable.

#### 5.4 XGBoost

```{r model-xgb, message=FALSE, warning=FALSE}
# TODO: [member name] — MODEL FITTING ONLY (rubric §5 = basic fit, tuning goes in §6.4)
# Use DEFAULT hyperparameters — do NOT tune here
# method = "xgbTree" | requires one-hot encoding via model.matrix(~ . - 1)
# CV: cv_ctrl | Weights: scale_pos_weight in xgb params
# Produce: confusionMatrix() on df_test, heatmap, 4 metrics
# Save metrics as: macro_f1_xgb, fatal_recall_xgb (needed for §6.4 before/after table)
# Save plots: p_xgb_cm, p_xgb_varimp
#   ggsave("plots/5_4_xgboost_confusion_matrix.png",          p_xgb_cm,     width=7, height=5, dpi=150)
#   ggsave("plots/5_4_xgboost_variable_importance_top15.png", p_xgb_varimp, width=7, height=5, dpi=150)

```

#### 5.5 K-Nearest Neighbours

```{r model-knn, message=FALSE, warning=FALSE}
# TODO: [member name] — MODEL FITTING ONLY (rubric §5 = basic fit, tuning goes in §6.5)
# Use DEFAULT k — do NOT tune here (default k = 5 or caret's single default)
# method = "kknn" | preProcess = c("center", "scale") | CV: cv_ctrl_smote
# Produce: confusionMatrix() on df_test, heatmap, 4 metrics
# Save metrics as: macro_f1_knn, fatal_recall_knn (needed for §6.5 before/after table)
# Save plots: p_knn_cm
#   ggsave("plots/5_5_knn_confusion_matrix.png", p_knn_cm, width=7, height=5, dpi=150)

```

#### 5.6 Naive Bayes

```{r model-nb, message=FALSE, warning=FALSE}
# TODO: [member name] — MODEL FITTING ONLY (rubric §5 = basic fit, tuning goes in §6.6)
# Use DEFAULT hyperparameters (usekernel = FALSE, laplace = 0) — do NOT tune here
# method = "naive_bayes" | CV: cv_ctrl_smote | Weights: via SMOTE + class prior
# Produce: confusionMatrix() on df_test, heatmap, 4 metrics
# Save metrics as: macro_f1_nb, fatal_recall_nb (needed for §6.6 before/after table)
# Save plots: p_nb_cm
#   ggsave("plots/5_6_naive_bayes_confusion_matrix.png", p_nb_cm, width=7, height=5, dpi=150)

```

------------------------------------------------------------------------

### 6 Model Improvement

#### 6.1 Multinomial Logistic Regression - Improvement

```{r improve-lr, message=FALSE, warning=FALSE}
# TODO: [member name] — MODEL IMPROVEMENT (rubric §6 = tuning/improvement, basic fit is in §5.1)
# Tune: decay via tuneGrid = expand.grid(decay = c(0, 0.001, 0.01, 0.1))
# Plot: decay vs Mean_F1 curve from model_lr_tuned$results
# Table: before/after using macro_f1_lr / fatal_recall_lr from §5.1 vs tuned values
```

#### 6.2 Decision Tree - Improvement

```{r improve-dt, message=FALSE, warning=FALSE}
# TODO: [member name] — MODEL IMPROVEMENT (rubric §6 = tuning/improvement, basic fit is in §5.2)
# Tune: cp via tuneGrid = expand.grid(cp = seq(0.0001, 0.01, by = 0.001))
# Plot: cp vs Mean_F1 curve from model_dt_tuned$results (or plotcp on standalone rpart fit)
# Table: before/after using macro_f1_dt / fatal_recall_dt from §5.2 vs tuned values
```

#### 6.3 Random Forest - Improvement

The primary hyperparameter for Random Forest is `mtry` — the number of candidate variables randomly sampled at each split. The default value of $\lfloor\sqrt{p}\rfloor$ is not necessarily optimal, especially for imbalanced multi-class problems where Fatal examples are sparse. `tuneRF()` was used to search over a sequence of `mtry` values via OOB error, using a 20,000-row subsample of the training features to keep search time tractable (OOB rankings stabilise at far smaller sample sizes than the full dataset).

```{r improve-rf, message=FALSE, warning=FALSE, fig.show='hold', out.width='33%'}
# Use a 20k-row subsample of rf_x_train to search mtry quickly via tuneRF().
# Finding the best mtry doesn't need the full dataset — OOB rankings are stable
# on a representative subsample. Final model is still fit on the full training set.
set.seed(42)
tune_idx <- sample(nrow(rf_x_train), min(nrow(rf_x_train), 20000))
invisible(capture.output(
  tune_result <- tuneRF(
    x          = rf_x_train[tune_idx, ],
    y          = df_train_sub$`Degree of crash`[tune_idx],
    mtryStart  = floor(sqrt(ncol(rf_x_train))),
    ntreeTry   = 50,
    stepFactor = 1.5,
    improve    = 0.01,
    trace      = FALSE,
    plot       = FALSE
  )
))
best_mtry <- tune_result[which.min(tune_result[, "OOBError"]), "mtry"]
cat("Best mtry from tuneRF:", best_mtry, "\n")

# mtry vs OOB error curve
p_rf_tune <- as.data.frame(tune_result) %>%
  ggplot(aes(x = mtry, y = OOBError)) +
  geom_line(colour = "#2E74B5", linewidth = 1) +
  geom_point(size = 3, colour = "#2E74B5") +
  geom_vline(xintercept = best_mtry, linetype = "dashed", colour = "#C0392B") +
  annotate("text", x = best_mtry, y = max(tune_result[, "OOBError"]),
           label = paste("Best mtry =", best_mtry),
           hjust = -0.1, colour = "#C0392B", size = 3.5) +
  labs(title = "Random Forest — Tuning: mtry vs OOB Error (20k subsample)",
       x = "mtry (variables randomly sampled per split)", y = "OOB Error Rate") +
  theme_minimal(base_size = 12)
print(p_rf_tune)
ggsave("plots/6_3_random_forest_mtry_tuning_curve.png", p_rf_tune, width = 7, height = 5, dpi = 150)

# Final tuned model fit — using df_train_sub (50%), swap to df_train for final submission
set.seed(42)
model_rf_tuned <- randomForest(
  x          = rf_x_train,
  y          = df_train_sub$`Degree of crash`,
  mtry       = best_mtry,
  ntree      = 100,
  importance = TRUE
)

pred_rf_tuned      <- predict(model_rf_tuned, rf_x_test)
cm_rf_tuned        <- confusionMatrix(pred_rf_tuned, df_test$`Degree of crash`)
fatal_recall_tuned <- cm_rf_tuned$byClass["Class: Fatal", "Recall"]
macro_f1_tuned     <- mean(cm_rf_tuned$byClass[, "F1"], na.rm = TRUE)
weighted_f1_tuned  <- weighted.mean(cm_rf_tuned$byClass[, "F1"],
                       prop.table(table(df_test$`Degree of crash`)), na.rm = TRUE)
kappa_tuned        <- cm_rf_tuned$overall["Kappa"]

cat(sprintf(
  "    RF Tuned  —  Test Set Metrics \nOOB error rate : %.4f\nMacro-F1       : %.4f\nFatal Recall   : %.4f\nWeighted F1    : %.4f\nKappa          : %.4f\n",
  model_rf_tuned$err.rate[100, "OOB"], macro_f1_tuned, fatal_recall_tuned, weighted_f1_tuned, kappa_tuned
))

# Before / After comparison table
data.frame(
  Model          = c("RF : default mtry", "RF : tuned mtry"),
  mtry           = c(floor(sqrt(ncol(rf_x_train))), best_mtry),
  `Macro-F1`     = round(c(macro_f1_rf,    macro_f1_tuned),    4),
  `Fatal Recall` = round(c(fatal_recall_rf, fatal_recall_tuned), 4),
  check.names = FALSE
) %>% knitr::kable(caption = "Random Forest — Before vs After Tuning")

# Plot 2: OOB error vs ntree convergence curve
p_rf_oob <- data.frame(ntree = seq_len(nrow(model_rf_tuned$err.rate)),
           OOB   = model_rf_tuned$err.rate[, "OOB"]) %>%
  ggplot(aes(x = ntree, y = OOB)) +
  geom_line(colour = "#2E74B5") +
  labs(title = "Random Forest : OOB Error vs Number of Trees (tuned mtry)",
       x = "Number of trees", y = "OOB error rate") +
  theme_minimal(base_size = 12)
print(p_rf_oob)
ggsave("plots/6_3_random_forest_oob_convergence.png", p_rf_oob, width = 7, height = 5, dpi = 150)

# Plot 3: Confusion matrix heatmap — tuned model (last in row)
p_rf_tuned_cm <- as.data.frame(cm_rf_tuned$table) %>%
  ggplot(aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = Freq), size = 6, fontface = "bold") +
  scale_fill_gradient(low = "white", high = "#1F4E79") +
  labs(title = "Random Forest (Tuned) — Confusion Matrix (Test Set)",
       subtitle = paste0("Fatal Recall: ", round(fatal_recall_tuned, 3),
                         "  |  Macro-F1: ", round(macro_f1_tuned, 3))) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
print(p_rf_tuned_cm)
ggsave("plots/6_3_random_forest_tuned_confusion_matrix.png", p_rf_tuned_cm, width = 7, height = 5, dpi = 150)
```

`tuneRF` identified **mtry = 13** as optimal, more than double the default of 6. Retraining with this value yielded consistent improvements across every metric: OOB error dropped from 18.4% to 18.1%, Macro-F1 increased from 0.614 to 0.634, Fatal Recall improved from 0.144 to 0.181, Weighted F1 from 0.813 to 0.815, and Kappa from 0.583 to 0.590. The OOB convergence curve confirms that 100 trees is sufficient — the OOB error stabilises before 75 trees, indicating no benefit from growing additional trees. The improvement in Fatal Recall (+0.037) is particularly meaningful: a larger `mtry` forces each tree to consider more variables per split, which helps the ensemble find the rare feature combinations that characterise fatal crashes rather than defaulting to the majority Injury signal. The confusion matrix shows that the tuned model correctly identified 68 fatal crashes (up from 54 with the default `mtry`), with the majority of misclassifications falling into Injury (302) rather than Non-casualty (6) — a safer failure mode, as fatal crashes are at least flagged as high-severity events.

#### 6.4 XGBoost - Improvement

```{r improve-xgb, message=FALSE, warning=FALSE}
# TODO: [member name] — MODEL IMPROVEMENT (rubric §6 = tuning/improvement, basic fit is in §5.4)
# Tune: nrounds, max_depth, eta via expand.grid(nrounds=c(50,100,200), max_depth=c(3,6), eta=c(0.01,0.1,0.3))
# Plot: best nrounds vs Mean_F1 curve from model_xgb_tuned$results
# Table: before/after using macro_f1_xgb / fatal_recall_xgb from §5.4 vs tuned values
```

#### 6.5 K-Nearest Neighbours - Improvement

```{r improve-knn, message=FALSE, warning=FALSE}
# TODO: [member name] — MODEL IMPROVEMENT (rubric §6 = tuning/improvement, basic fit is in §5.5)
# Tune: kmax via tuneGrid = expand.grid(kmax = c(5, 11, 21, 31, 51), distance = 2, kernel = "optimal")
# Plot: k vs Mean_F1 curve from model_knn_tuned$results
# Table: before/after using macro_f1_knn / fatal_recall_knn from §5.5 vs tuned values
```

#### 6.6 Naive Bayes - Improvement

```{r improve-nb, message=FALSE, warning=FALSE}
# TODO: [member name] — MODEL IMPROVEMENT (rubric §6 = tuning/improvement, basic fit is in §5.6)
# Tune: usekernel and laplace via tuneGrid = expand.grid(usekernel = c(TRUE, FALSE), laplace = c(0, 0.5, 1), adjust = 1)
# Plot: bar chart of Macro-F1 across tuning combinations
# Table: before/after using macro_f1_nb / fatal_recall_nb from §5.6 vs tuned values
```

------------------------------------------------------------------------

### 7 Model Evaluation & Selection

<!-- TODO: [member name] — comparison table (6 models × 4 metrics), best model selection + justification -->

------------------------------------------------------------------------

### 8 Final Results & Conclusion

<!-- TODO: [member name] — answer RQ, top predictors, practical implications, process reflection -->

------------------------------------------------------------------------

### References

<!-- TODO: [member name] — APA 7th format references for all sources cited -->

------------------------------------------------------------------------

### Acknowledgement of AI

<!-- TODO: [member name] — AI acknowledgement statement -->
