---
title: ""
author: ""
output:
  html_document:
    code_folding: hide
---

```{r library, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(readr)
library(forcats)
library(scales)
library(knitr)
library(caret)
library(nnet)
library(rpart)
library(rpart.plot)
library(randomForest)
library(xgboost)
library(kknn)
library(e1071)
library(themis)
library(pROC)
library(flextable)
```

```{css, echo=FALSE}
body { font-family: "Times New Roman", Times, serif; font-size: 10pt; line-height: 1.3; }
h2 { margin-top: 0.4em; margin-bottom: 0.15em; }
h3 { margin-top: 0.3em; margin-bottom: 0.1em; }
h4 { margin-top: 0.25em; margin-bottom: 0.1em; }
p  { margin-top: 0.2em; margin-bottom: 0.2em; }
h1.title { display: none; }
h4.author { display: none; }
h4.date   { display: none; }
```

```{r load-data, message=FALSE, warning=FALSE, results='hide'}

```

### 1 Research Question (Recap)

Given environmental conditions (weather, lighting, road surface, speed limit) and traffic unit characteristics (TU type, manoeuvre, number of units), can we predict NSW road crash severity (Fatal / Injury / Non-casualty): a three-class classification problem?

---

### 2 Preprocessing

All data-dependent steps are performed **within cross-validation folds** to prevent data leakage (James et al., 2021).

#### 2.1 Step 1 — Data Integration & TU-Level Feature Aggregation

```{r step1-integration, message=FALSE, warning=FALSE, results='hide'}

```

#### 2.2 Step 2 — Feature Removal (Leakage, Identifiers, Sparse Columns)

```{r step2-removal, message=FALSE, warning=FALSE, results='hide'}

```

#### 2.3 Step 3 — Missing Value Imputation

```{r step3-imputation, message=FALSE, warning=FALSE, results='hide'}

```

#### 2.4 Step 4 — Feature Engineering

```{r step4-engineering, message=FALSE, warning=FALSE, results='hide'}

```

#### 2.5 Step 5 — Outlier Detection

```{r step5-outliers, message=FALSE, warning=FALSE, results='hide'}

```

#### 2.6 Step 6 — Encoding & LGA Pooling

```{r step6-encoding, message=FALSE, warning=FALSE, results='hide'}

```

#### 2.7 Step 7 — Train/Test Split (80/20 Stratified)

```{r step7-split, message=FALSE, warning=FALSE, results='hide'}

```

#### 2.8 Step 8 — Class Imbalance: SMOTE

```{r step8-smote, message=FALSE, warning=FALSE, results='hide'}

```

#### 2.9 Step 9 — CV Setup (Nested 5-Fold)

```{r step9-cv, message=FALSE, warning=FALSE, results='hide'}

```

---

### 3 Models

#### 3.1 Multinomial Logistic Regression

```{r model-lr, message=FALSE, warning=FALSE}

```

#### 3.2 Decision Tree

```{r model-dt, message=FALSE, warning=FALSE}

```

#### 3.3 Random Forest

```{r model-rf, message=FALSE, warning=FALSE}

```

#### 3.4 XGBoost

```{r model-xgb, message=FALSE, warning=FALSE}

```

#### 3.5 KNN

```{r model-knn, message=FALSE, warning=FALSE}

```

#### 3.6 Naive Bayes

```{r model-nb, message=FALSE, warning=FALSE}

```

---

### 4 Evaluation & Results

#### 4.1 Per-Model Confusion Matrices

```{r eval-confusion, message=FALSE, warning=FALSE}

```

#### 4.2 Metric Comparison Table

```{r eval-metrics-table, message=FALSE, warning=FALSE}

```

#### 4.3 Variable Importance (Random Forest & XGBoost)

```{r eval-varimp, message=FALSE, warning=FALSE}

```

---

### 5 Conclusion

---

### References

Chawla, N. V., Bowyer, K. W., Hall, L. O., & Kegelmeyer, W. P. (2002). SMOTE: Synthetic minority over-sampling technique. *Journal of Artificial Intelligence Research*, *16*, 321–357. https://doi.org/10.1613/jair.953

James, G., Witten, D., Hastie, T., & Tibshirani, R. (2021). *An introduction to statistical learning: With applications in R* (2nd ed.). Springer. https://doi.org/10.1007/978-1-0716-1418-1

Transport for NSW. (2024). *NSW road crash data 2020–2024* [Dataset]. Australian Government Open Data Platform. https://www.data.gov.au/data/dataset/nsw-2-nsw-crash-data

Transport for NSW. (2025). *NSW road safety progress report 2024*. Centre for Road Safety. https://www.transport.nsw.gov.au/system/files/media/documents/2025/crs_road_safety_progress_report_2024.pdf

### Acknowledgement of AI

Group 07 (Workshop 12) declares that all ideas, analysis, and written content presented in this report are our own. We did not use generative AI to write our code. We used ChatGPT (GPT-5.3, OpenAI, https://chatgpt.com) and Claude Sonnet 4.6 (Anthropic, https://claude.ai) for code refinement, grammar checking, and reviewing our report against the marking rubric.
