---
title: "NSW Road Crash Severity Prediction"
subtitle: "STAT5003 — Group 07, Workshop 12"
author: "Sanskar | Robert | Haipeng | Chen | Yihan"
date: "21 May 2026"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.align = "center", out.width = "90%")

library(ggplot2)
library(dplyr)
library(readr)
library(forcats)
library(scales)
library(knitr)
library(kableExtra)
library(tidyr)
library(flextable)
```

```{r load-data, include=FALSE}
df_raw <- read_csv("nsw_road_crash_merged_2020_2024.csv", show_col_types = FALSE)

df_raw <- df_raw %>%
  rename(severity = `Degree of crash`) %>%
  mutate(severity = factor(severity,
    levels = c("Fatal", "Injury", "Non-casualty (towaway)"),
    labels = c("Fatal", "Injury", "Non-casualty")))
```

## Research Question & Motivation

**Research Question:**

> Given environmental conditions (weather, lighting, road surface, speed limit) 
> and traffic unit characteristics (TU type, manoeuvre, number of units), 
> can we predict NSW road crash severity **(Fatal / Injury / Non-casualty)**?
> — A three-class classification problem.

**Motivation:**

- NSW recorded **340 road fatalities in 2024**, over two-thirds in regional areas *(TfNSW, 2025)*
- Fatal crashes represent only **~1.4%** of all crashes → severe class imbalance
- A predictive model has direct value for:

| Stakeholder | Value |
|-------------|-------|
| **Traffic Authorities** | Targeted speed and road interventions |
| **Emergency Responders** | Faster, evidence-based resource allocation |
| **Policymakers & Insurers** | Precise policy and pricing decisions |

## EDA & Data Understanding: Dataset Overview

```{r dataset-overview-robert, echo=FALSE, message=FALSE, warning=FALSE}
total_records <- nrow(df_raw)
total_vars    <- ncol(df_raw)
class_counts  <- table(df_raw$severity)
class_props   <- round(prop.table(class_counts) * 100, 1)

overview <- data.frame(
  Item = c(
    "Time Period",
    "Data Sources",
    "Merge Key",
    "Total Records",
    "Total Variables",
    "Target Variable",
    "Fatal",
    "Injury",
    "Non-casualty"
  ),
  Detail = c(
    "2020 – 2024",
    "CRASH.xlsx + TRAFFIC UNIT.xlsx (TfNSW via data.gov.au)",
    "Crash ID (left join)",
    format(total_records, big.mark = ","),
    as.character(total_vars),
    "Degree of Crash (3-class classification)",
    paste0(format(class_counts["Fatal"], big.mark=","), 
           " (", class_props["Fatal"], "%)"),
    paste0(format(class_counts["Injury"], big.mark=","), 
           " (", class_props["Injury"], "%)"),
    paste0(format(class_counts["Non-casualty"], big.mark=","), 
           " (", class_props["Non-casualty"], "%)")
  )
)

flextable(overview) |>
  set_header_labels(Item = "Item", Detail = "Detail") |>
  theme_vanilla() |>
  bold(j = 1) |>
  bold(part = "header") |>
  bg(i = c(2, 4, 6), bg = "#f5f5f5") |>
  color(i = 7, j = 2, color = "#d62728") |>
  color(i = 8, j = 2, color = "#ff7f0e") |>
  color(i = 9, j = 2, color = "#2ca02c") |>
  fontsize(size = 16, part = "all") |>    
  fontsize(size = 18, part = "header") |> 
  padding(padding = 8, part = "all") |> 
  set_table_properties(width = 1) |>   
  autofit()
```

## EDA & Data Understanding: Class Imbalance

```{r class-imbalance-robert, echo=FALSE, message=FALSE, warning=FALSE, fig.width=7, fig.height=4}

class_dist <- df_raw %>%
  count(severity) %>%
  mutate(
    proportion = round(n / sum(n) * 100, 1),
    label = paste0(proportion, "%\n(n=", format(n, big.mark=","), ")")
  )

ggplot(class_dist,
       aes(x = severity, y = proportion, fill = severity)) +
  geom_bar(stat = "identity", width = 0.6) +
  geom_text(
    aes(label = label),
    vjust = -0.3,
    size  = 4,
    fontface = "bold"
  ) +
  scale_fill_manual(values = c(
    "Fatal"        = "#d62728",
    "Injury"       = "#ff7f0e",
    "Non-casualty" = "#2ca02c"
  )) +
  scale_y_continuous(limits = c(0, 80)) +
  labs(
    title    = "Class Distribution of Crash Severity (2020–2024)",
    subtitle = "Fatal crashes are severely underrepresented",
    x = "Severity Class",
    y = "Proportion (%)"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

>  **Fatal crashes (~1.4%) are severely underrepresented**  
>  SMOTE applied **inside training fold only**  
>  Primary metrics: **Macro-F1** and **Fatal Recall (≥ 0.50)**
## EDA & Data Understanding: Dataset Overview

<small style="color:#2E74B5">**Rubric criterion:** EDA & Data Understanding  \|  2 pts</small>

```{r slide-dataset}
# TODO: [member name] — dataset summary table (source, records, variables, target, date range)
```

## EDA & Data Understanding: Speed vs Severity

<small style="color:#2E74B5">**Rubric criterion:** EDA & Data Understanding  \|  2 pts</small>

```{r slide-speed, fig.height=3.5}
speed_levels <- c("10 km/h","20 km/h","30 km/h","40 km/h","50 km/h",
                  "60 km/h","70 km/h","80 km/h","90 km/h","100 km/h","110 km/h")

df_raw %>%
  filter(`Speed limit` %in% speed_levels) %>%
  mutate(`Speed limit` = factor(`Speed limit`, levels = speed_levels)) %>%
  count(`Speed limit`, severity) %>%
  group_by(`Speed limit`) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = `Speed limit`, y = prop, fill = severity)) +
  geom_col(position = "stack") +
  geom_hline(yintercept = 0.014, linetype = "dashed", colour = "white", linewidth = 0.8) +
  scale_fill_manual(values = c("Fatal" = "#d62728", "Injury" = "#ff7f0e", "Non-casualty" = "#2ca02c")) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(title = "Crash Severity Proportion by Speed Limit",
       subtitle = "Fatal proportion rises sharply at 100-110 km/h  |  dashed line = overall Fatal baseline (~1.4%)",
       x = "Speed Limit", y = "Proportion of Crashes", fill = "Severity") +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "right")
```

> **Key finding:** Fatal proportion doubles at 100 km/h and triples at 110 km/h relative to 50 km/h zones — directly motivates `speed_band` feature engineering.

## EDA & Data Understanding: High-Risk Scenarios

<small style="color:#2E74B5">**Rubric criterion:** EDA & Data Understanding  \|  2 pts</small>

```{r slide-highrisk, fig.height=3.8}
risk_table <- df_raw %>%
  filter(!is.na(`Speed limit`), !is.na(`Natural lighting`), !is.na(Urbanisation)) %>%
  group_by(`Speed limit`, `Natural lighting`, Urbanisation) %>%
  summarise(
    `Total Crashes` = n(),
    `Fatal %`       = mean(severity == "Fatal"),
    .groups = "drop"
  ) %>%
  filter(`Total Crashes` >= 200) %>%
  arrange(desc(`Fatal %`)) %>%
  slice_head(n = 10) %>%
  mutate(`Fatal %` = scales::percent(`Fatal %`, accuracy = 0.1)) %>%
  rename(`Speed Limit` = `Speed limit`, `Lighting` = `Natural lighting`)

kable(risk_table, align = c("l","l","l","r","r")) %>%
  kable_styling(full_width = TRUE, font_size = 13) %>%
  row_spec(0, bold = TRUE, background = "#2E74B5", color = "white") %>%
  row_spec(1:3, background = "#fce4e4") %>%
  column_spec(5, bold = TRUE, color = "#d62728")
```

> **Key finding:** Highest-risk scenarios all share 110 km/h + darkness + rural conditions — three features directly captured by `speed_band`, `is_night`, and `Urbanisation`.

## EDA & Data Understanding: Temporal Patterns

<small style="color:#2E74B5">**Rubric criterion:** EDA & Data Understanding  \|  2 pts</small>

```{r slide-temporal, fig.height=3.5}
day_levels <- c("Monday","Tuesday","Wednesday","Thursday","Friday","Saturday","Sunday")

df_raw %>%
  mutate(`Day of week of crash` = factor(`Day of week of crash`, levels = day_levels),
         `Two-hour intervals`   = factor(`Two-hour intervals`,
                                          levels = unique(`Two-hour intervals`))) %>%
  group_by(`Day of week of crash`, `Two-hour intervals`) %>%
  summarise(count = n(), .groups = "drop") %>%
  ggplot(aes(x = `Two-hour intervals`, y = `Day of week of crash`, fill = count)) +
  geom_tile(colour = "white", linewidth = 0.4) +
  scale_fill_gradient(low = "#f0f4ff", high = "#1F4E79", labels = comma) +
  labs(title = "Crash Frequency by Day of Week and Two-Hour Interval",
       subtitle = "Morning (7-9 am) and evening (4-6 pm) commute peaks visible on weekdays",
       x = "Two-Hour Interval", y = NULL, fill = "Crash Count") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank())
```

> **Key finding:** Volume peaks at weekday commute hours, motivating `is_rush_hour` and `is_weekend` features. Fatal *rate* (not count) is elevated on weekends and late-night hours.

## Data Preprocessing: Pipeline

<small style="color:#2E74B5">**Rubric criterion:** Data Preprocessing  \|  2 pts</small>

**Steps 1–5: Cleaning & Feature Engineering**

1.  **TU Integration** — aggregated traffic-unit flags per crash (`has_motorcycle`, `has_heavy_vehicle`, `has_pedestrian`, `has_cyclist`) using `group_by + summarise + left_join`
2.  **Feature removal** — dropped leakage columns (post-crash injury counts), identifiers, and variables with \>75% missing values; dropped `Degree of crash - detailed` (redundant with target)
3.  **Imputation** — removed rows \>50% missing; categorical → `"Unknown"`; numeric → median
4.  **Feature engineering** — `is_weekend`, `is_rush_hour`, `is_night`, `is_wet`, speed bands — all motivated by EDA findings
5.  **Outlier handling** — NSW geographic bounds filter; winsorise `Distance` at 99th percentile

**Steps 6–9: Modelling Setup**

6.  **Encoding** — trees use native R `factor`; KNN/LR use one-hot; rare LGA levels → `"Other"`
7.  **Scaling** — Z-score standardisation within fold for KNN; trees are scale-invariant
8.  **Class imbalance** — inverse-frequency class weights (all models); SMOTE applied **inside training fold only**
9.  **Split** — 2024 data sealed as temporal holdout; 80/20 stratified split on 2020–2023; 5-fold CV

## Data Preprocessing: Feature Engineering & Class Balancing

<small style="color:#2E74B5">**Rubric criterion:** Data Preprocessing  \|  2 pts</small>

**Engineered features — all motivated by EDA**

| Feature | Derivation | EDA motivation |
|------------------------|------------------------|------------------------|
| `speed_band` | Cut `Speed limit` into ≤50 / 60–80 / 90–100 / 110 | Fatal rate rises sharply above 100 km/h |
| `is_night` | `Natural lighting` ∈ {Darkness, Dawn, Dusk} | Dark conditions double fatal risk |
| `is_rush_hour` | `Two-hour intervals` ∈ {7–9, 16–18} | Commute peaks dominate crash volume |
| `is_weekend` | `Day of week of crash` ∈ Saturday, Sunday | Weekend crashes differ in severity profile |
| `is_wet` | `Surface condition` contains "Wet" | Wet roads linked to skidding fatalities |
| `has_motorcycle` | TU-level aggregated flag | Motorcyclists over-represented in fatals |
| `has_pedestrian` | TU-level aggregated flag | Pedestrian crashes nearly always severe |

**Handling class imbalance**

-   Raw Fatal share \~1.4% → accuracy-maximising models ignore Fatal entirely
-   Strategy: **inverse-frequency class weights** (all models) + **SMOTE** inside training fold (LR, KNN)
-   SMOTE raises synthetic Fatal share to \~10% in training only — test set remains unmodified

## Model Fitting: Approach & Setup

<small style="color:#2E74B5">**Rubric criterion:** Model Fitting  \|  2 pts</small>

```{r model-overview}
model_tbl <- tibble::tibble(
  Model              = c("Multinomial Logistic Regression", "Decision Tree",
                         "Random Forest", "XGBoost",
                         "K-Nearest Neighbours", "Naive Bayes"),
  `R package`        = c("nnet", "rpart", "randomForest", "xgboost", "kknn", "e1071"),
  `Why chosen`       = c(
    "Interpretable baseline; log-odds show predictor direction",
    "Fully interpretable decision rules; native categorical handling",
    "Ensemble; robust to noise; provides variable importance",
    "Gradient boosting; typically strongest on tabular data",
    "Non-parametric; captures local structure without assumptions",
    "Probabilistic baseline; reveals feature independence violations"
  ),
  `Imbalance strategy` = c("Class weights + SMOTE", "Class weights",
                            "Class weights", "scale_pos_weight",
                            "Class weights + SMOTE", "Class prior")
)
kable(model_tbl, booktabs = TRUE) %>%
  kable_styling(full_width = TRUE, font_size = 14) %>%
  column_spec(1, bold = TRUE, width = "6em") %>%
  column_spec(3, width = "18em") %>%
  row_spec(0, bold = TRUE, background = "#2E74B5", color = "white")
```

*All models trained via `caret` with 5-fold CV. Primary metric: **Macro-F1**. Hard constraint: **Fatal Recall ≥ 0.50**.*

## Model Fitting: Logistic Regression

<small style="color:#2E74B5">**Rubric criterion:** Model Fitting  \|  2 pts</small>

```{r results-lr, fig.height=3.5}
# TODO: [member name] — confusion matrix heatmap + coefficient plot or ROC curves
# Train on: df_train | CV: cv_ctrl_smote | Evaluate on: df_test
```

## Model Fitting: Decision Tree

<small style="color:#2E74B5">**Rubric criterion:** Model Fitting  \|  2 pts</small>

```{r results-dt, fig.height=3.5}
# TODO: [member name] — rpart.plot tree diagram + confusion matrix heatmap
# Train on: df_train | CV: cv_ctrl | Evaluate on: df_test
```

## Model Fitting: Random Forest

<small style="color:#2E74B5">**Rubric criterion:** Model Fitting  \|  2 pts</small>

```{r results-rf, fig.height=3.5, out.width='48%', fig.show='hold'}
knitr::include_graphics(c(
  "plots/5_3_random_forest_variable_importance_top15.png",
  "plots/5_3_random_forest_confusion_matrix.png"
))
```

<div style="font-size:13px">
| Metric | Value |
|--------|-------|
| OOB Error Rate | 18.4% |
| Macro-F1 | 0.614 |
| Fatal Recall | 0.144 |
| Weighted F1 | 0.813 |
| Kappa | 0.583 |
</div>

> **Top predictors:** Street lighting, LGA, Day of week, Month of crash, Key TU type — environmental context dominates over vehicle type.

## Model Fitting: XGBoost

<small style="color:#2E74B5">**Rubric criterion:** Model Fitting  \|  2 pts</small>

```{r results-xgb, fig.height=3.5}
# TODO: [member name] — variable importance (by Gain) + training loss curve + confusion matrix
# Train on: df_train | CV: cv_ctrl | Evaluate on: df_test
```

## Model Fitting: KNN & Naive Bayes

<small style="color:#2E74B5">**Rubric criterion:** Model Fitting  \|  2 pts</small>

```{r results-knn-nb, fig.height=3.5}
# TODO: [member name] — k vs Macro-F1 curve (KNN) + side-by-side confusion matrices
# KNN: cv_ctrl_smote | NB: cv_ctrl_smote | Evaluate both on: df_test
```

## Model Improvement: Hyperparameter Tuning

<small style="color:#2E74B5">**Rubric criterion:** Model Improvement  \|  2 pts</small>

```{r slide-improvement}
tuning_tbl <- tibble::tibble(
  Model       = c("Logistic Regression", "Decision Tree", "Random Forest",
                  "XGBoost", "KNN", "Naive Bayes"),
  Parameter   = c("decay", "cp (complexity)", "mtry",
                  "nrounds / max_depth / eta", "k (neighbours)", "usekernel / laplace"),
  `Search Range` = c("0, 0.001, 0.01, 0.1",
                     "0.0001 to 0.01",
                     "tuneRF: 4, 6, 9, 13",
                     "nrounds: 50-200 | depth: 3-6 | eta: 0.01-0.3",
                     "5, 11, 21, 31, 51",
                     "kernel: T/F | laplace: 0, 0.5, 1"),
  `Best Value`       = c("TODO", "TODO", "mtry = 13",  "TODO", "TODO", "TODO"),
  `Macro-F1 Before`  = c("TODO", "TODO", "0.614",      "TODO", "TODO", "TODO"),
  `Macro-F1 After`   = c("TODO", "TODO", "0.634",      "TODO", "TODO", "TODO"),
  `Fatal Recall After` = c("TODO", "TODO", "0.181",    "TODO", "TODO", "TODO")
)

kable(tuning_tbl, booktabs = TRUE) %>%
  kable_styling(full_width = TRUE, font_size = 12) %>%
  row_spec(0, bold = TRUE, background = "#2E74B5", color = "white") %>%
  row_spec(3, background = "#eaf3fb") %>%
  column_spec(1, bold = TRUE)
```

> RF: `tuneRF` identified **mtry = 13** (default = 6) — Macro-F1 +0.020, Fatal Recall +0.037. OOB converges before 75 trees.

## Model Evaluation

<small style="color:#2E74B5">**Rubric criterion:** Model Evaluation  \|  2 pts</small>

```{r slide-evaluation}
eval_tbl <- tibble::tibble(
  Model          = c("Logistic Regression", "Decision Tree", "Random Forest (tuned)",
                     "XGBoost", "KNN", "Naive Bayes"),
  `Macro-F1`     = c("TODO", "TODO", "**0.634**", "TODO", "TODO", "TODO"),
  `Fatal Recall` = c("TODO", "TODO", "0.181",     "TODO", "TODO", "TODO"),
  `Weighted F1`  = c("TODO", "TODO", "0.815",     "TODO", "TODO", "TODO"),
  `Kappa`        = c("TODO", "TODO", "0.590",     "TODO", "TODO", "TODO"),
  `Meets Fatal Recall >= 0.50` = c("TODO", "TODO", "No (dev subsample)", "TODO", "TODO", "TODO")
)

kable(eval_tbl, booktabs = TRUE, escape = FALSE) %>%
  kable_styling(full_width = TRUE, font_size = 13) %>%
  row_spec(0, bold = TRUE, background = "#2E74B5", color = "white") %>%
  row_spec(3, background = "#eaf3fb") %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(3, color = ifelse(eval_tbl$`Fatal Recall` == "TODO", "grey50", "#d62728"))
```

> RF (tuned) leads on Macro-F1 among completed models. Fatal Recall below 0.50 threshold on 50% subsample — expected to improve on full training data for final submission.

## Model Selection

<small style="color:#2E74B5">**Rubric criterion:** Model Selection  \|  2 pts</small>

```{r slide-selection}
# TODO: [member name] — best model name, justification, top predictors from variable importance
```

## Final Results & Conclusion

<small style="color:#2E74B5">**Rubric criterion:** Final Results & Conclusion  \|  2 pts</small>

```{r slide-conclusion}
# TODO: [member name] — answer to RQ, key findings, practical implications for stakeholders
```

## Reflection & Future Directions

<small style="color:#2E74B5">**Rubric criterion:** Final Results & Conclusion  \|  2 pts (continued)</small>

```{r slide-reflection}
# TODO: [member name] — limitations, future research directions
```

