---
title: ""
author: ""
output:
  html_document:
    code_folding: hide
---

```{r library, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(readr)
library(forcats)
library(scales)
library(knitr)
library(caret)
library(nnet)
library(rpart)
library(rpart.plot)
library(randomForest)
library(xgboost)
library(kknn)
library(e1071)
library(themis)
library(pROC)
library(flextable)
```

```{css, echo=FALSE}
body { font-family: "Times New Roman", Times, serif; font-size: 10pt; line-height: 1.3; }
h2 { margin-top: 0.4em; margin-bottom: 0.15em; }
h3 { margin-top: 0.3em; margin-bottom: 0.1em; }
h4 { margin-top: 0.25em; margin-bottom: 0.1em; }
p  { margin-top: 0.2em; margin-bottom: 0.2em; }
h1.title { display: none; }
h4.author { display: none; }
h4.date   { display: none; }
```

```{r load-data, message=FALSE, warning=FALSE, results='hide'}
df <- read.csv("nsw_road_crash_merged_2020_2024.csv", check.names = FALSE)
cat("Dimensions:", nrow(df), "rows x", ncol(df), "cols\n")
str(df)
```

### 1 Research Question (Recap)

Given environmental conditions (weather, lighting, road surface, speed limit) and traffic unit characteristics (TU type, manoeuvre, number of units), can we predict NSW road crash severity (Fatal / Injury / Non-casualty): a three-class classification problem?

---

### 2 Preprocessing

All data-dependent steps are performed **within cross-validation folds** to prevent data leakage (James et al., 2021).

#### 2.1 Step 1 — Data Integration & TU-Level Feature Aggregation

```{r step1-integration, message=FALSE, warning=FALSE, results='hide'}
#Each row is one traffic unit — summarise binary flags per crash, then join back
tu_flags <- df %>%
  group_by(`Crash ID`) %>%
  summarise(
    has_motorcycle    = as.integer(any(`TU type group` == "Motorcycle")),
    has_heavy_vehicle = as.integer(any(`TU type group` %in% c("Articulated truck", "Heavy rigid truck", "Bus"))),
    has_pedestrian    = as.integer(any(`TU type group` == "Pedestrian")),
    has_cyclist       = as.integer(any(`TU type group` == "Pedal cycle")),
    tu_count          = n(),
    .groups = "drop"
  )

df <- df %>% left_join(tu_flags, by = "Crash ID")

cat("TU flags added. Rows:", nrow(df), "| New cols:", ncol(df), "\n")
```

#### 2.2 Step 2 — Feature Removal (Leakage, Identifiers, Sparse Columns)

```{r step2-removal, message=FALSE, warning=FALSE, results='hide'}
#Leakage: post-crash outcome counts that directly encode the target
leakage_cols <- c("No. killed", "No. seriously injured",
                  "No. moderately injured", "No. minor-other injured")

# Redundant: detailed severity is a finer version of the target column
# "Degree of crash" is already the aggregated 3-class target we are predicting
redundant_target_cols <- c("Degree of crash - detailed")

# Identifiers and free-text location fields (no predictive signal)
id_cols <- c("Crash ID", "Traffic unit ID", "Street of crash",
             "Street of travel", "Town", "Identifying feature",
             "Identifying feature type", "Conurbation 1")

# High-NA columns (>75% missing — confirmed from inspection)
# Primary hazardous feature 98.7%, Primary temporary feature 98.5%,
# Object hit 2 97.1%, DCA supplement 91.6%, Object hit 1 85.2%,
# Primary permanent feature 75.1%
high_na_cols <- c("Primary hazardous feature", "Primary temporary feature",
                  "Object hit 2", "DCA supplement", "Object hit 1",
                  "Primary permanent feature")

# Route no. (36.3% NA, redundant with location fields)
redundant_cols <- c("Route no.", "RUM - code", "RUM - description",
                    "DCA - code", "DCA - description")

drop_cols <- unique(c(leakage_cols, redundant_target_cols, id_cols, high_na_cols, redundant_cols))
df_clean  <- df %>% select(-any_of(drop_cols))

cat("Columns after removal:", ncol(df_clean), "\n")
```

#### 2.3 Step 3 — Missing Value Imputation

```{r step3-imputation, message=FALSE, warning=FALSE, results='hide'}
#Only 3 NAs in Identifying feature (already dropped)
# Other TU type: 17.1% NA — treat as "Unknown" (not present)
# Remaining categoricals: fill "Unknown" for any residual NAs

df_clean <- df_clean %>%
  mutate(
    `Other TU type` = ifelse(is.na(`Other TU type`), "Unknown", `Other TU type`),
    across(where(is.character), ~ replace_na(., "Unknown")),
    across(where(is.numeric),   ~ replace_na(., median(., na.rm = TRUE)))
  )

cat("Missing values after imputation:", sum(is.na(df_clean)), "\n")
```

#### 2.4 Step 4 — Feature Engineering

```{r step4-engineering, message=FALSE, warning=FALSE, results='hide'}
#Two-hour intervals are stored as strings e.g. "8:00 AM - 9:59 AM"
# Extract start hour for time-based features
df_clean <- df_clean %>%
  mutate(
    crash_hour   = as.integer(sub(":.*", "", `Two-hour intervals`)),
    is_weekend   = as.integer(`Day of week of crash` %in% c("Saturday", "Sunday")),
    is_rush_hour = as.integer(crash_hour %in% c(7, 8, 9, 16, 17, 18)),
    is_night     = as.integer(`Natural lighting` %in% c("Darkness", "Dawn", "Dusk")),
    is_wet       = as.integer(`Surface condition` %in% c("Wet", "Snow or ice")),
    speed_band   = case_when(
      `Speed limit` %in% c("10 km/h","20 km/h","30 km/h","40 km/h","50 km/h") ~ "low_50",
      `Speed limit` %in% c("60 km/h","70 km/h","80 km/h")                     ~ "mid_60_80",
      `Speed limit` %in% c("90 km/h","100 km/h")                              ~ "high_90_100",
      `Speed limit` == "110 km/h"                                              ~ "very_high_110",
      TRUE                                                                     ~ "unknown"
    )
  )

cat("Engineered features added. Columns:", ncol(df_clean), "\n")
```

#### 2.5 Step 5 — Outlier Detection

```{r step5-outliers, message=FALSE, warning=FALSE, results='hide'}
#Distance: winsorise at 99th percentile
p99_dist <- quantile(df_clean$Distance, 0.99, na.rm = TRUE)
df_clean <- df_clean %>%
  mutate(Distance = pmin(Distance, p99_dist))

# Latitude/Longitude: drop records outside NSW bounds
# NSW approx: lat -28.2 to -37.5, lon 141.0 to 153.6
df_clean <- df_clean %>%
  filter(is.na(Latitude)  | (Latitude  >= -37.5 & Latitude  <= -28.2),
         is.na(Longitude) | (Longitude >= 141.0  & Longitude <= 153.6))

cat("Rows after outlier removal:", nrow(df_clean), "\n")
```

#### 2.6 Step 6 — Encoding & LGA Pooling

```{r step6-encoding, message=FALSE, warning=FALSE, results='hide'}
#Pool rare LGA values (<1% of records) into "Other"
lga_freq  <- prop.table(table(df_clean$LGA))
rare_lgas <- names(lga_freq[lga_freq < 0.01])
df_clean  <- df_clean %>%
  mutate(LGA = ifelse(LGA %in% rare_lgas, "Other", LGA))

# Convert all character columns to factors (required by caret/rpart/randomForest)
# Target: ordered factor with Fatal first for metric calculations
df_model <- df_clean %>%
  mutate(across(where(is.character), as.factor)) %>%
  mutate(`Degree of crash` = factor(`Degree of crash`,
           levels = c("Fatal", "Injury", "Non-casualty (towaway)")))

cat("Class distribution:\n")
print(prop.table(table(df_model$`Degree of crash`)))
cat("LGA levels after pooling:", nlevels(df_model$LGA), "\n")
```

#### 2.7 Step 7 — Train/Test Split (80/20 Stratified)

```{r step7-split, message=FALSE, warning=FALSE, results='hide'}
# Temporal holdout: 2024 data sealed off before any train/test split
# Only used once at the very end to check temporal generalisation
df_holdout_2024 <- df_model %>% filter(`Reporting year` == 2024)
df_model_2020_23 <- df_model %>% filter(`Reporting year` != 2024)

cat("Holdout (2024):", nrow(df_holdout_2024), "rows\n")
cat("Modelling set (2020-2023):", nrow(df_model_2020_23), "rows\n")

# 80/20 stratified split on 2020-2023 data only
set.seed(42)
train_idx <- createDataPartition(df_model_2020_23$`Degree of crash`, p = 0.8, list = FALSE)
df_train  <- df_model_2020_23[ train_idx, ]
df_test   <- df_model_2020_23[-train_idx, ]

cat("Train:", nrow(df_train), "| Test:", nrow(df_test), "\n")
cat("Train class distribution:\n")
print(prop.table(table(df_train$`Degree of crash`)))
```

#### 2.8 Step 8 — Class Imbalance: SMOTE

```{r step8-smote, message=FALSE, warning=FALSE, results='hide'}
#Class weights for models that accept them (LR, DT, RF)
# Inverse frequency weighting so Fatal gets highest weight
class_weights <- 1 / prop.table(table(df_train$`Degree of crash`))
class_weights <- class_weights / sum(class_weights)
cat("Class weights:\n"); print(round(class_weights, 4))

# SMOTE is applied inside caret's trainControl via the sampling argument
# where supported (see CV setup below) — not applied globally here
# to avoid leakage between folds
```

#### 2.9 Step 9 — CV Setup (Nested 5-Fold)

```{r step9-cv, message=FALSE, warning=FALSE, results='hide'}
#Standard 5-fold CV control — used by all models
cv_ctrl <- trainControl(
  method          = "cv",
  number          = 5,
  classProbs      = TRUE,
  summaryFunction = multiClassSummary,
  savePredictions = "final"
)

# CV control with SMOTE — used for models where class weights are not available
cv_ctrl_smote <- trainControl(
  method          = "cv",
  number          = 5,
  classProbs      = TRUE,
  summaryFunction = multiClassSummary,
  savePredictions = "final",
  sampling        = "smote"
)

cat("CV controls ready.\n")
```

---

### 3 Models

#### 3.1 Multinomial Logistic Regression

```{r model-lr, message=FALSE, warning=FALSE}
# Encoding & scaling: add preProcess = "nzv" in train()
# method = "multinom" handles factors via dummy variables internally
# Train on: df_train | CV control: cv_ctrl | Weights: class_weights
# Evaluate on: df_test | Temporal check: df_holdout_2024

```

#### 3.2 Decision Tree

```{r model-dt, message=FALSE, warning=FALSE}
# Encoding & scaling: none required — rpart handles R factors natively
# Train on: df_train | CV control: cv_ctrl | Weights: class_weights
# Evaluate on: df_test | Temporal check: df_holdout_2024

```

#### 3.3 Random Forest

```{r model-rf, message=FALSE, warning=FALSE}
# Encoding & scaling: none required — randomForest handles R factors natively
# Train on: df_train | CV control: cv_ctrl | Weights: classwt = class_weights
# Evaluate on: df_test | Temporal check: df_holdout_2024

```

#### 3.4 XGBoost

```{r model-xgb, message=FALSE, warning=FALSE}
# Encoding: requires numeric matrix — use model.matrix(~ . - 1, data = df_train) to one-hot encode
# Train on: df_train | CV control: cv_ctrl | Weights: scale_pos_weight
# Evaluate on: df_test | Temporal check: df_holdout_2024

```

#### 3.5 KNN

```{r model-knn, message=FALSE, warning=FALSE}
# Encoding & scaling: add preProcess = c("center", "scale") in train()
# This z-scores numerics and dummy-encodes factors automatically
# Train on: df_train | CV control: cv_ctrl_smote | Weights: via SMOTE
# Evaluate on: df_test | Temporal check: df_holdout_2024

```

#### 3.6 Naive Bayes

```{r model-nb, message=FALSE, warning=FALSE}
# Encoding & scaling: none required — naive_bayes handles factors directly
# Train on: df_train | CV control: cv_ctrl_smote | Weights: via SMOTE
# Evaluate on: df_test | Temporal check: df_holdout_2024

```

---

### 4 Evaluation & Results

#### 4.1 Per-Model Confusion Matrices

```{r eval-confusion, message=FALSE, warning=FALSE}

```

#### 4.2 Metric Comparison Table

```{r eval-metrics-table, message=FALSE, warning=FALSE}

```

#### 4.3 Variable Importance (Random Forest & XGBoost)

```{r eval-varimp, message=FALSE, warning=FALSE}

```

---

### 5 Conclusion

---

### References

Chawla, N. V., Bowyer, K. W., Hall, L. O., & Kegelmeyer, W. P. (2002). SMOTE: Synthetic minority over-sampling technique. *Journal of Artificial Intelligence Research*, *16*, 321–357. https://doi.org/10.1613/jair.953

James, G., Witten, D., Hastie, T., & Tibshirani, R. (2021). *An introduction to statistical learning: With applications in R* (2nd ed.). Springer. https://doi.org/10.1007/978-1-0716-1418-1

Transport for NSW. (2024). *NSW road crash data 2020–2024* [Dataset]. Australian Government Open Data Platform. https://www.data.gov.au/data/dataset/nsw-2-nsw-crash-data

Transport for NSW. (2025). *NSW road safety progress report 2024*. Centre for Road Safety. https://www.transport.nsw.gov.au/system/files/media/documents/2025/crs_road_safety_progress_report_2024.pdf

### Acknowledgement of AI

Group 07 (Workshop 12) declares that all ideas, analysis, and written content presented in this report are our own. We did not use generative AI to write our code. We used ChatGPT (GPT-5.3, OpenAI, https://chatgpt.com) and Claude Sonnet 4.6 (Anthropic, https://claude.ai) for code refinement, grammar checking, and reviewing our report against the marking rubric.
